\documentclass{wg21}

\usepackage{xcolor}
\usepackage{soul}
\usepackage{ulem}
\usepackage{fullpage}
\usepackage{parskip}
\usepackage{csquotes}
\usepackage{relsize}      % provide relative font size changes
\usepackage{textcomp}     % provide \text{l,r}angle
\usepackage{mathrsfs}     % mathscr font

\title{Movability of Single-pass Iterators}
\docnumber{D1207R2}
\audience{LWG}
\author{Corentin Jabot}{corentin.jabot@gmail.com}

\usepackage[inner=1cm,outer=2cm]{geometry}

\begin{document}
\maketitle

\begin{flushright}
\hfill \break
\hfill \break
\textit{I want to move(it), move(it), y'all want to move(it);}
\end{flushright}

\hypertarget{changes}{%
	\section{Changes}\label{changes}}

\hypertarget{revision-2}{%
	\subsection{Revision 2}\label{revision-2}}

\begin{itemize}
	\item
	Remove discussions about \texttt{ContiguousIterator}
\end{itemize}

\hypertarget{revision-1}{%
	\subsection{Revision 1}\label{revision-1}}

\begin{itemize}
	\item
	Refine the impact on the standard
	\item
	Mention \texttt{ITER\_CONCEPT}
	\item
	Remove the idea the \texttt{ranges::copy} could move from non-copyable
	iterators
	\item
	Replace Cpp17Iterator by LegacyIterator to reflect the standard
	\item
	Remove the very wrong idea of having view return \texttt{begin()} by
	reference.
	\item
	Fix some confusing phrasing
\end{itemize}

\hypertarget{introduction}{%
	\section{Introduction}\label{introduction}}

Non-forward Input iterators and output iterators, also known as
``Single-pass iterators'' are semantically move-only. The standard
states:

Note: For input iterators, a == b does not imply ++a == ++b (Equality
does not guarantee the substitution property or referential
transparency.) Algorithms on input iterators should never attempt to
pass through the same iterator twice. They should be single pass
algorithms.

This means that once an iterator is copied, only one of the copies can
be read-from if either one is incremented, which make the usefulness of
such object questionable. Deferencing multiple copies of a single pass
iterator often exposes undefined or invalid behavior if either one is
incremented: the following example exposes Undefined behavior: 
\begin{codeblock}

auto other = some_input_iterator;
std::cout << *(++other) << *some_input_iterator << '\n';
\end{codeblock}

It would, therefore, make sense that classes satisfying the
\texttt{InputIterator} concept shall only be required to be movable.

Alas, Single-pass iterators and many classes satisfying its requirements
predate C++11, they do therefore have move only semantic with copy
syntax. In that regard, they are similar to \texttt{auto\_ptr}.

\hypertarget{terminology}{%
	\section{Terminology}\label{terminology}}

This paper redefines the requirements of some concepts as specified in
the Working Draft In the rest of this paper

\begin{itemize}
	\item
	\texttt{InputIterator} designs the \texttt{InputIterator} concept as
	proposed by this paper
	\item
	\texttt{Cpp20InputIterator} designs the \texttt{InputIterator} concept
	as specified in the Working Draft.
	\item
	\texttt{OutputIterator} designs the \texttt{OutputIterator} concept as
	proposed by this paper
	\item
	\texttt{OutputIterator} designs the \texttt{OutputIterator} concept as
	proposed in the Working Draft
\end{itemize}

\hypertarget{scope}{%
	\section{Scope}\label{scope}}

This paper proposes changes targeting C++20. Because the modifications
proposed here changes some requirements and concepts introduced by
Ranges, the authors strongly suggest they are considered for the
inclusion in the same version of the standard. Indeed, Concepts
introduced by ranges gives us a unique opportunity to make the
modifications proposed, as they might, in some cases, break code, if
introduced after the publication of C++20.

\hypertarget{non-goal}{%
	\subsection{Non-Goal}\label{non-goal}}

As a large amount of code depends on the Input/Output iterators
requirements as specified by C++17, this paper does not propose any
modifications to the \texttt{LegacyInputIterator} or any class that
depends on it. Specifically, we do not propose to change the
requirements or wording of \texttt{istream\_iterator},
\texttt{ostream\_iterator}, \texttt{istreambuf\_iterator} or
\texttt{ostreambuf\_iterator}. Furthermore, we do not propose
modifications to algorithms in the namespace \texttt{std}. The new
iterators we propose here are in fact mostly incompatible with existing
algorithms. They are meant to be used in the \texttt{ranges} namespace
and as basic building blocks of range-based views.

While the ability to use move-only iterators with the algorithms defined
in the \texttt{std} namespace would certainly be welcomed, doing so
would weaken the \texttt{Cpp20InputIterator} concept and leads to other
issues (namely, \texttt{std} based algorithms require iterators to be
\texttt{EqualityComparable}, which the \texttt{Cpp20InputIterator} does
not require).

In practice, that means that types satisfying the
\texttt{LegacyInputIterator} requirements continue to work unaffected
with algorithms defined in the \texttt{std} namespace. They may not be
compatible with algorithms defined in the ranges namespace, or with new
code using non-movable types satisfying the \texttt{InputIterator}
concept as proposed here.

Inversely, types satisfying the \texttt{InputIterator} concepts may not
be compatible with algorithms in \texttt{std} as they may not be able to
satisfy the \texttt{LegacyInputIterator} requirements \textbf{if} they
are not copyable.

Because it hardly makes sense to copy an Input Iterator (more on that
later), it would be possible to add support for move-only iterators to
the \texttt{std} namespace without much change to the standard. However,
because implementers may copy iterators within the implementation of the
standard library, along with existing third-party libraries, a lot of
code would need to be adapted. And there is little pressure to do so as
existing iterators types cannot be changed.

\textbf{Furthermore, while we propose to add support for movable
	non-forward iterators, the proposed design does not preclude, in any
	way, the existance of copyable non-forward iterators.}

\hypertarget{motivation}{%
	\section{Motivation}\label{motivation}}

\hypertarget{move-only-state}{%
	\subsection{Move-only state}\label{move-only-state}}

It may be desirable for an iterator to hold a move-only object, becoming
itself move-only, which is not possible with iterators modeling
LegacyIterator. A real-world example of such iterator is described in
\cite{P0902}. While syntactically copyable in the current design,
a \texttt{coroutine\_handle} such as used by a \texttt{generator} input
iterator ought to be move-only.

\hypertarget{implicitly-destructive-operations}{%
	\subsection{Implicitly destructive
		operations}\label{implicitly-destructive-operations}}

Reading from an input sequence is a destructive operation. But that
destruction is reflected nowhere in the API. Less experienced developers
may not be aware of the destructive / single-pass nature of non-forward
Iterators By making \texttt{InputIterator} move only, developers will
have to explicitly move them, which both signals the invalidation of the
move-from object, but, more importantly, that the underlying data will
be destroyed.

\hypertarget{what-is-a-move-only-iterator}{%
	\section{What is a move-only
		iterator?}\label{what-is-a-move-only-iterator}}

Unlike \cite{P0902}, we do not propose to introduce a new
iterator category.

A move-only Iterator is a non-forward iterator (either input or output
depending on whether is it writable). This means that a move-only
iterator has \emph{almost} the same semantic requirements as an
\texttt{InputIterator}, and offers the same operations. In other words,
everything that can be expressed and done with a
\texttt{Cpp20InputIterator} can be equally expressed and done with a
move-only/non-copyable \texttt{InputIterator}.

Therefore, this paper does not propose to introduce a new iterator
category, new named-requirement, concept name or iterator tag.

Furthermore, there is no \texttt{ForwardIterator} that is only movable,
as a \texttt{ForwardIterator} is by definition an iterator that can be
copied. We will expand on this later.

\hypertarget{a-holistic-approach-to-iterators}{%
	\section{A Holistic Approach to
		Iterators}\label{a-holistic-approach-to-iterators}}

While the first part of this paper focuses on making move-only iterators
possible, as a means to get some code to compile, it is important to
take a step back and to think about what movability means for Iterators,
from first principles.

An iterator denotes a position into a sequence of elements (whether that
sequence maps to memory or not is, for our purpose, irrelevant).

A most basic iterator can be incremented, which means it can move to the
next position in the sequence. An iterator does not own the sequence
iterated over (there are exceptions, ie: generators), which means the
salient property of an iterator is its position in that sequence.

Iterators categories then represent the way an iterator can move along
that sequence. 
\begin{itemize}
	\item Input and FordwardIterator: sequentially, one direction
	\item BidirectionalIterator: sequentially, both directions
	\item RandomAccess: both directions in O(1)
\end{itemize}

ContiguousIterator is an optimization of RandomAccessIterator specific
to the C++ memory model that further, constrain the underlying sequence
to be laid out contiguously in memory.

Stepanov theorized an additional category, ``Index iterator'', which has
O(1) access but in a single direction.

Further work was made on iterator categories, notably the Boost.Iterator
library focused on separating traversal (how the iterator moves along
the sequence) from access (whether dereferencing an iterator allows the
pointed element to be read, written or both). While a very interesting
concept, it falls outside the scope of this paper. Just keep in mind
that everything that applies to non-forward \texttt{InputIterator}
usually applies to OutputIterator - which are always non-Forward, the
standard lacking that symmetry between read access and write access.

However, focusing on traversal, the set of iterators categories is
actually rather closed, there are only so many ways a sequence can be
traversed. An important point of Stepanov design is that each category
is a refinement of the preceding one. \texttt{RandomAccessIterator} is a
\texttt{BidirectionalIterator} which in turn is a
\texttt{ForwardIterator}. Every algorithm applicable to a
\texttt{ForwardIterator} can be equally applied to a
\texttt{BidirectionalIterator}, etc.

So, what separates \texttt{InputIterator} from \texttt{ForwardIterator}
if they are both ``forward'' in that they can both traverse a sequence
in one direction?

\texttt{ForwardIterator} is defined as being ``multi-pass''. Meaning it
can traverse a sequence multiple times. That, in turn, implies
\texttt{ForwardIterator} is copyable, because if a sequence can be
traversed multiple times, it can also be traversed multiple times at the
same time and therefore there can be multiple \texttt{ForwardIterator}
pointing at different elements in the sequence. ForwardIterator is also
always \texttt{EqualityComparable}. Two \texttt{ForwardIterator} compare
equal if they point to the same elements in the sequence (remember, that
in the general case, the position of an iterator in a sequence is its
sole salient property). And so \texttt{ForwardIterator}, being both
\texttt{EqualityComparable} and \texttt{Copyable} is Regular.

The standard defines the ``multi pass'' guarantee by stating:
\textgreater{} a == b implies ++a == ++b \textgreater{} Given X is a
pointer type or the expression (void)++X(a), \emph{a is equivalent to
	the expression }a.

In other words: Two identical objects to which is applied the same
transformation are identical. 

Copying a \texttt{FordwardIterator} copies
the salient properties of that value and incrementing it does not modify
the underlying sequence. So \texttt{ForwardIterator} is required to be a
regular type behaving like a regular type.

Which bring us to \texttt{InputIterator}. InputIterator is a ``single
pass'' iterator. The underlying sequence can on only be traversed once.
The existence of an Iterator at the nth position in the sequence implies
there can be no valid iterator at the position n-1 in that same
sequence.

\begin{codeblock}
//Given an InputIterator a 
auto b = a; a++;
b; // is invalid.
\end{codeblock}

However, remember that the sole salient property of an iterator is its
distance to the start of the sequence. Incrementing an iterator only
mutates that property (again, conceptually, independently of
implementation). And the only operation that mutates that property is
the increment operation (which Stepanov calls \texttt{successor}).

This implies that as a non-forward iterator moves from one element of
the sequence to the next, that element is destroyed.

All of this is well known and is basically rephrasing ``Input iterators
are single pass''.

An important point to make is that how an iterator can traverse a
sequence is derived from the nature of the sequence rather than from the
iterator itself. The point could be made that there is no such thing as
an ``Input iterator'' Or a ``Forward Iterator'' because what we really
mean is ``Iterator over an Input Sequence'' or ``Iterator over a Forward
Sequence''.

This is saying that, to be able to reason properly about iterators and
traversal, we must assume that the iterator type associated with a
sequence is the most specialized possible for that sequence.

The problem is, of course, that we do not have, in the general case, a
more meaningful way to express the traversability of a sequence than by
defining what type of iterator is used to iterate over it.

It is then the responsibility of the developer providing the sequence to
define the most appropriate -- the most specialized -- iterator category
for that sequence.

In practice, because \texttt{InputIterator} and \texttt{ForwardIterator}
are syntactically identical and because of the single-pass /
multi-passes guarantees are poorly taught, it is common for iterators to
be mis-categorized. Other iterator categories do not have these problems
as each subsequent refining category adds syntax requirements:
BidirectionalIiterator require decrement operators, RandomAccessIterator
has further requirements.

But then, is there a set of operations and semantic requirements,
translating to actual C++ syntax, that could allow for InputIterator to
be easily distinguished from each other? Can we avoid requiring a tag
system? Is there a defining operation that distinguishes
\texttt{InputIterator} from \texttt{ForwardIterator} in such a way that
it would both not require an explicit category tagging while at the same
time offering a better understanding of iterator categories as well as a
less surprising and safer API for non-forward iterators?

In fact, there is. We established that \texttt{ForwardIterator}s are
semantically copyable, while \texttt{InputIterator}s are not. So the
requirement that promotes an \texttt{InputIterator} into a
\texttt{ForwardIterator} is indeed copyability - which translate in C++
to a copy constructor. We can, therefore, consider that, in the absence
of a tag, all non-copyable iterators are \texttt{InputIterator}, while
all copyable iterators are \texttt{ForwardIterator}.

This model, however, deviates slightly from Stepanov's work and
\texttt{LegacyInputIterator}: 

Copying a \texttt{LegacyInputIterator}
does not invalidate either copy. In fact, it is quite valid to deference
multiple copies of a \texttt{LegacyInputIterator}.

Elements Of Programming has the notion of \texttt{Regular} types (and in
Stepanov's work all Iterators are regular), but also the notion of
regular transformations (aka pure functions) - which, given the same
input, always give the same output. Given a \texttt{ForwardIterator}
\texttt{fi}, there is a \texttt{successor} function returning an
incremented copy of \texttt{fi} such as
\texttt{sucessor(fi)\ ==\ sucessor(fi)}. In C++, that regular
\texttt{sucessor} function is
\texttt{ForwardIterator::operator++(int);}, in that
\texttt{(it++)\ ==\ (it++)} for any given \texttt{ForwardIterator}.

For \texttt{InputIterator}, Stepanov specifies that the
\texttt{successor} is a pseudo transformation or a non-regular
transformation that look like a regular one. And therein lies the rub.

Like a pointer, \texttt{InputIterator} is Regular, up until the point a
transformation of an instance affects all copies.

\begin{codeblock}
	InputIterator i = /*...*/
	*i    //ok
	auto a = i //ok
	*i    //ok
	i++;  // a now invalid
\end{codeblock}

This design accurately models the nature of iterators. Because an
iterator represents a position in a sequence, it is natural that
multiple iterators could point to the same position. After one copy is
incremented, in Stepanov's model, other copies are in a partially formed
state and cannot be used (but they can be assigned to, or destroyed).

Let's consider the case where we move from an iterator instead of
copying it.


\begin{codeblock}
	InputIterator i = /*...*/
	*i           //ok
	auto a = move(i); //ok
	*i;          //invalid
	a++;         //ok
	i++;         //invalid
\end{codeblock}


Moving from an iterator invalidates it early, albeit artificially. As
per standard, the moved-from iterator is in a valid, but unspecified
state, and cannot be used (but can be assigned to, or destroyed). Notice
the similarity between ``a valid, but unspecified state'' and ``a
partially formed state''.

The difference is slim. Notably, both models are equally expressive.
References can be used, should multiple names be necessary. In
Stepanov's model iterators are made invalid by the natural mutation of
the sequence upon increment rather than by artificially preventing
multiple copies.

The second model in which the iterator is moved from, the one we think
should be the default way to handle non-forward iterators, is however a
much better fit for the C++ model, and offers much stronger guarantees
to both the human developer as well as static analysis tools.

In the ``increment invalidates'' model, objects are spiritually
moved-from at a distance, which neither the theory of special relativity
nor the C++ memory, model are equipped to handle. This makes it hard for
tools to detect invalid uses - although it might become possible with
better tools (See Herb Sutter's CppCon2018 talk). But most concerning,
there is no way for a developer to know that the iterators are
entangled.

\begin{codeblock}
	auto i = troubles.begin();
	auto schrodingers_iterator = i;
	i++;
	auto nasal_demon = *schrodingers_iterator;
\end{codeblock}

The code above might be perfectly fine. Indeed whether it is well
defined or not depends on whether the iterator return by
\texttt{troubles.begin();} is forward or not. It is undecidable in these
4 lines of slide-code. It is not much more obvious in a complex program
that may pass iterators to other functions or store them in containers,
etc. There are, after all, no theoretical limits to the distance in time
and space over which entanglement perdures.

Even worse, should the type of \texttt{troubles.begin();} be changed
from Forward to Input, the code would change from perfectly fine to UB,
with no warning.

Moving non-forward iterators, therefore, better expresses intent, is
safer and less surprising. Move-only non-forward Iterators also express
the destructive nature of incrementation and give a better sense of the
difference between \texttt{InputIterator} and \texttt{ForwardIterator}.

\hypertarget{an-holistic-approach-to-iterator-tags-and-iterator-concepts}{%
	\subsection{An Holistic Approach to Iterator Tags and Iterator
		Concepts}\label{an-holistic-approach-to-iterator-tags-and-iterator-concepts}}

Missing the notion of movability pre-c++11 and lacking concepts,
\texttt{LegacyIterator}s are syntactically distinguished by tags. a
\texttt{LegacyInputIterator} is one which has an
\texttt{input\_iterator\_tag} tag, while a
\texttt{LegacyForwardIterator} is one which has a
\texttt{forward\_iterator\_tag} tag. This creates a sort of circular,
self-referential definition. This has carried over to the \texttt{Iterator} concepts definitions. 

Iterators concepts then :

\begin{itemize}
	\item Have semantic requirements not expressed through syntax and therefore not enforceable at compile time
	\item Need syntax to artificially subscribe to the correct, most refined concept
\end{itemize}


Of course, it is not always possible to express all of a type's semantic
requirements through syntax, and in some cases, tags are an unfortunate
necessity. However, they should be the mechanism of last recourse, and
whenever possible, the semantic requirements should be reflected in the
syntax. The idea is that hidden requirements not expressed as code lead
to easier-to-misuse types, which inevitably translates to runtime bugs.
\textbf{Ultimately, requirements that can neither be checked at compile
	time (concepts) or runtime (contracts) are bound to be ignored}. Rooted
in the belief that not all birds quack like a duck, this proposal
leverages meaningful syntactic requirements to increase the type safety
of the iterator taxonomy.

In the case of iterators, all requirements of all iterators categories
can be expressed syntactically:

\begin{codeblock}
	template <class I> concept bool InputIterator =
	Readable<I> &&
	Iterator<I> ;
	
	template <class I> concept bool ForwardIterator =
	InputIterator<I> &&
	Copyable<I> &&
	EqualityComparable<I>;
	
	template <class I> concept bool BidirectionalIterator =
	ForwardIterator<I> &&
	Decrementable<I>;
	
	template <class I> concept bool RandomAccessIterator =
	BidirectionalIterator<I> &&
	RandomAccessIncrementable<I>;
\end{codeblock}


This is of course simplified but shows that each iterator category
subsumes the last and adds a single, cohesive set of requirement
enforceable at compile-time. In this design, there is no risk of a type
satisfying the wrong concept because of a poorly chosen tag.

\hypertarget{tags-as-an-opt-in-opt-out-mechanism}{%
	\subsubsection{Tags as an opt-in opt-out
		mechanism}\label{tags-as-an-opt-in-opt-out-mechanism}}

Iterators concepts already support semantic-only checking of iterator
requirements for types that do not define either iterator\_category or
iterator concept. Currently, this machinery will identify categories
from ForwardIterator to RandomAccessIterator. With this proposal,
non-copyable tagless types that otherwise meet the requirements of
InputIterator be correctly identified as non-forward InputIterator,
which is always the correct assumption. Copyable tagless iterators will
remain categorized as ForwardIterator by that machinery.

\hypertarget{qa}{%
	\section{Q/A}\label{qa}}

\hypertarget{non-regular-iterators-really}{%
	\subsection{Non-regular iterators,
		really?}\label{non-regular-iterators-really}}

This proposal advocates for Non-Regular Iterators, and weakens
\texttt{WeaklyIncrementable} requirements to that effect. Non-Regularity
is best avoided, so this might feel like going backward.

However, \textbf{non-regular types are easier to reason about than types
	that just pretend to be regular}. Because \texttt{InputIterator} is
meant to iterate over a non-regular sequence, it is not regular (whether
we like it or not), and the best we can do is make sure the syntax
matches the semantic. It would be accurate to say that
\texttt{InputIterator} is locally regular, but this doesn't help much in
the context of the c++ memory model. This paper is in part motivated by
the conviction that exposing \textbf{a false sense of (Semi-)regularity
	is much more detrimental to code robustness than non-regularity}.

\hypertarget{what-about-equality-of-input-iterators}{%
	\subsection{What about Equality of Input
		Iterators?}\label{what-about-equality-of-input-iterators}}

A first, misguided, version of this paper attempted to prevent
comparability of types meeting the \texttt{InputIterator} requirements.
\texttt{InputIterator} should, in general, not be
\texttt{EqualityComparable}, since they cannot be copied and a
fundamental idea in Stepanov's teachings is that copy and equality are
two sides of the same coin.

However, preventing \texttt{Equality} requires dramatic changes to the
design and the author was reminded that negative-requirements are in
general a terrible idea.

Early feedback suggested a desire to be able to compare non-forward
iterators. Consider the following:

\begin{codeblock}
auto a = stream.begin(); 
auto b = stream.begin();
if(a == b) {
}
\end{codeblock}

This code will inevitably lead to suffering at some point. However, we
cannot prevent people from constructing multiple non-forward iterators,
and these iterators will compare equal until one of them invalidate the
other.

Two non-forward iterators compare equal if-and-only-if they point to the
same position of the same sequence (and only one such position can be
referred to at any given time).

Allowing \texttt{EqualityComparable} on non-forward iterators also
simplify the interoperability of \texttt{std::} and \texttt{ranges::}
iterators. However, the author would like to recommend that all future
non-forward iterators introduced in the standard be \emph{not}
\texttt{EqualityComparable}. Instead, non-forward iterator should
compare to a Sentinel, which is a much better model.
\texttt{common\_iterator} can be used to ease migration and
interoperability.

\hypertarget{but-moved-from-objects-are-still-objects}{%
	\subsection{But\ldots{} Moved-from objects are still
		objects!}\label{but-moved-from-objects-are-still-objects}}

Sure, moving-from leaves a trail of objects in an unspecified state.
However, it is much more easy for tools and humans alike to understand
that moved-from objects should not be used, and in fact, all majors
compilers can warn about these patterns. We think that for the case at
hand, focusing on the proper handling of values -- as opposed to objects
-- is a sufficient approximation to reduce the potential for iterators
misuse while not weakening the stronger mathematical underpinning of the
STL.

\hypertarget{does-iterators-default-constructability-needs-revisiting}{%
	\subsection{Does iterators default-constructability needs
		revisiting?}\label{does-iterators-default-constructability-needs-revisiting}}

Default-constructability of iterator seems to have been added, removed
and added back to the Ranges TS and the One Ranges Proposal several
times. To the best of my knowledge, this was done for the sake of
Semiregularity. Given that this proposal strikes semi-regularity, should
this question be revisited?

The authors want to point out that default-constructed iterators are
almost never in a specified state and are almost always unsafe to use.
Moreover, DefaultConstructible is not a requirement of any algorithm
using ranges and ultimately, we think enforcing DefaultConstructibility
weakens the better \texttt{Sentinel} model introduced by ranges.

\hypertarget{what-about-p0902r0}{%
	\subsection{What about \cite{P0902}?}\label{what-about-p0902r0}}

Andrew Hunter's ``Move-only iterators'' paper proposes a design to
introduce Move-Only iterators in the taxonomy of
\texttt{LegacyIterator}. However, this design does not offer a solution
to use these move-only iterators with existing algorithms, limiting
their usefulness. The iterators proposed by P0902 are additionally
\texttt{EqualityComparable}. The advantage of that is that they are
compatible with algorithms designed with C++17 downward. That's,
however, a potential source of bugs and confusion.

However, if LEWG feels strongly about a solution compatible with
existing algorithms it would be possible to relax the requirements of
concerned algorithms to accept move-only iterators. along with the
introduction of a new \texttt{move\_iterator\_tag} trait.

Such algorithms would then be compatible with types satisfying
\texttt{InputIterator} (as proposed by this paper) through a
\texttt{common\_iterator} adaptor.

If proven with enough confidence that requirements of existing
algorithms in the \texttt{std} namespace can be relaxed to handle
move-only iterator, the necessary modifications can be applied in a
subsequent standard version.

So while there would definitively be value in supporting move-only
iterators everywhere it makes sense, and the potential for breakage is
relatively low, we do not propose it for lack of visibility on the
consequences of such changes.

\hypertarget{why-do-you-want-to-take-my-copyable-inputiterators-away-from-me-i-like-them}{%
	\subsection{Why do you want to take my Copyable InputIterators away from
		me, I like
		them?!}\label{why-do-you-want-to-take-my-copyable-inputiterators-away-from-me-i-like-them}}

We do not propose anything of the sort. But, we propose that
\begin{itemize}
	\item Any \texttt{InputIterator} that happens to be \texttt{Copyable} is also a \texttt{ForwardIterator}.
	\item  It remains possible to opt-out of that behavior by defining \texttt{iterator\_concept} to be
	\texttt{input\_iterator\_tag}.
\end{itemize}

\lstnewenvironment{tabularlstlisting}[1][]
{ \lstset{basicstyle=\tiny}{#1} }
{}

\begin{tabular}{|l|l|l|}
Non-copyable Iterator & Copyable Iterator & Copyable Iterator with a tag\\
\hline
\begin{tabularlstlisting}
struct It {
  It(It&&) = default;
  It(const It&) = delete;
  //...
};
\end{tabularlstlisting} & \begin{tabularlstlisting}
struct It {
  It(const It&) = default;
  //...
};
\end{tabularlstlisting} &
\begin{tabularlstlisting}
struct It {
  It(const It&) = default;
  using iterator_concept = input_iterator_tag;
  //...
};
\end{tabularlstlisting}
\\
\hline
\begin{tabularlstlisting}
static_assert(InputIterator<It>);
static_assert(!ForwardIterator<It>);
\end{tabularlstlisting} & 
\begin{tabularlstlisting}
static_assert(InputIterator<It>);
static_assert(ForwardIterator<It>);
\end{tabularlstlisting} &
\begin{tabularlstlisting}
static_assert(InputIterator<It>);
static_assert(!ForwardIterator<It>);
\end{tabularlstlisting}
\\
\end{tabular}


\hypertarget{will-this-break-existing-code}{%
	\subsection{Will this break existing code
		?!}\label{will-this-break-existing-code}}

We want to reiterate(!) that all the changes proposed in this paper are
only applicable to concepts, types, and requirements that were added to
the standard by the Ranges proposal. They do not, in any way, impact
code depending on types, requirements or algorithms as defined by the
C++17 standard

\hypertarget{wont-that-implicit-categorization-lead-to-miss-categorization}{%
	\subsection{Won't that implicit categorization lead to
		miss-categorization?}\label{wont-that-implicit-categorization-lead-to-miss-categorization}}

The only valid use cases for \texttt{InputIterator} are streams or other
input devices, and iterators that own a non-copyable generator. Most
views and iterators are Forward. It turns out that C++ types are
\texttt{Copyable} by default, therefore, Iterators will be categorized
as \texttt{ForwardIterator} by default, which is correct in most cases.

This proposal is also a teaching opportunity because the nature of
\texttt{InputIterator} is often poorly understood and misconstrued. We
suspect that these tweaks to the taxonomy of Iterator will make them
easier to teach.

\hypertarget{post-increment-on-non-copyable-iterators}{%
	\subsection{Post Increment on non-copyable
		iterators}\label{post-increment-on-non-copyable-iterators}}

Post-incrementing move-only iterators would obviously be incorrect.
However, a satisfying solution was offered by \cite{P0541}


\hypertarget{implementation-experience}{%
	\section{Implementation experience}\label{implementation-experience}}

We validated the design in cmstl2. However, cmcstl2 deviates from the
Working Draft as it doesn't have the same Deep Integration system and
therefore lacks the \texttt{ITER\_CONCEPT} machinery. Furthermore, we
have not yet completed this work. Some algorithms, like \texttt{count},
proved to require specialization for \texttt{InputIterator} because of
implementation specific details.

\hypertarget{acknowledgments}{%
	\subsection{Acknowledgments}\label{acknowledgments}}

The authors like to thank Connor Waters, Tony Van Eerd, Eric Niebler,
Casey Carter, Christopher Di Bella, Sean Parent and Arthur O'Dwyer who
gave tremendously helpful feedbacks during the writing of this paper.


\section{Wording}


\rSec2[iterator.concepts]{Iterator concepts}


\rSec3[iterator.concept.winc]{Concept \tcode{WeaklyIncrementable}}

\pnum
The \tcode{WeaklyIncrementable} concept specifies the requirements on
types that can be incremented with the pre- and post-increment operators.
The increment operations are not required to be equality-preserving,
nor is the type required to be \libconcept{EqualityComparable}.

\indexlibrary{\idxcode{WeaklyIncrementable}}%
\begin{codeblock}
	template<class I>
	concept WeaklyIncrementable =
	@\changed{Semiregular}{Movable}@<I> &&
	requires(I i) {
		typename iter_difference_t<I>;
		requires SignedIntegral<iter_difference_t<I>>;
		{ ++i } -> Same<I&>; // not required to be equality-preserving
		i++; // not required to be equality-preserving
	};
\end{codeblock}

\pnum
Let \tcode{i} be an object of type \tcode{I}. When \tcode{i} is in the domain of
both pre- and post-increment, \tcode{i} is said to be \term{incrementable}.
\tcode{I} models \tcode{WeaklyIncrementable<I>} only if

\begin{itemize}
	\item The expressions \tcode{++i} and \tcode{i++} have the same domain.
	\item If \tcode{i} is incrementable, then both \tcode{++i}
	and \tcode{i++} advance \tcode{i} to the next element.
	\item If \tcode{i} is incrementable, then
	\tcode{addressof(++i)} is equal to
	\tcode{addressof(i)}.
\end{itemize}

\pnum
\begin{note}
	For \added{\tcode{CopyConstructible}} \tcode{WeaklyIncrementable} types, \tcode{a} equals \tcode{b} does not imply that \tcode{++a}
	equals \tcode{++b}. (Equality does not guarantee the substitution property or referential
	transparency.) Algorithms on weakly incrementable types should never attempt to pass
	through the same incrementable value twice. They should be single-pass algorithms. These algorithms
	can be used with istreams as the source of the input data through the \tcode{istream_iterator} class
	template.
\end{note}

\rSec3[iterator.concept.inc]{Concept \tcode{Incrementable}}

\pnum
The \tcode{Incrementable} concept specifies requirements on types that can be incremented with the pre-
and post-increment operators. The increment operations are required to be equality-preserving,
and the type is required to be \libconcept{EqualityComparable}.
\begin{note}
	This supersedes the annotations on the increment expressions
	in the definition of \tcode{WeaklyIncrementable}.
\end{note}

\indexlibrary{\idxcode{Incrementable}}%
\begin{codeblock}
	template<class I>
	concept Incrementable =
	Regular<I> &&
	WeaklyIncrementable<I> &&
	requires(I i) {
		{ i++ } -> Same<I>;
	};
\end{codeblock}

\pnum
Let \tcode{a} and \tcode{b} be incrementable objects of type \tcode{I}.
\tcode{I} models \libconcept{Incrementable} only if

\begin{itemize}
	\item If \tcode{bool(a == b)} then \tcode{bool(a++ == b)}.
	\item If \tcode{bool(a == b)} then \tcode{bool(((void)a++, a) == ++b)}.
\end{itemize}

\pnum
\begin{note}
	The requirement that
	\tcode{a} equals \tcode{b}
	implies
	\tcode{++a} equals \tcode{++b}
	(which is not true for weakly incrementable types)
	allows the use of multi-pass one-directional
	algorithms with types that model \libconcept{Increment\-able}.
\end{note}

\rSec3[iterator.concept.iterator]{Concept \tcode{Iterator}}

\pnum
The \libconcept{Iterator} concept forms the basis
of the iterator concept taxonomy; every iterator models \libconcept{Iterator}.
This concept specifies operations for dereferencing and incrementing
an iterator. Most algorithms will require additional operations
to compare iterators with sentinels\iref{iterator.concept.sentinel}, to
read\iref{iterator.concept.input} or write\iref{iterator.concept.output} values, or
to provide a richer set of iterator movements (\ref{iterator.concept.forward},
\ref{iterator.concept.bidir}, \ref{iterator.concept.random.access}).

\indexlibrary{\idxcode{Iterator}}%
\begin{codeblock}
	template<class I>
	concept Iterator =
	requires(I i) {
		{ *i } -> @\placeholder{can-reference}@;
	} &&
	WeaklyIncrementable<I>;
\end{codeblock}

\begin{addedm}
\begin{note}
	Unlike the \oldconcept{Iterator} requirements\iref{input.iterators},
	the \libconcept{Iterator} concept does requires copyability of single-pass iterators.
\end{note}
\end{addedm}

\rSec3[iterator.concept.forward]{Concept \tcode{ForwardIterator}}

\pnum
The \libconcept{ForwardIterator} concept adds \added{copyability,} equality comparison and
the multi-pass guarantee, specified below.

\indexlibrary{\idxcode{ForwardIterator}}%
\begin{codeblock}
	template<class I>
	concept ForwardIterator =
	InputIterator<I> &&
	DerivedFrom<@\placeholdernc{ITER_CONCEPT}@(I), forward_iterator_tag> &&
	Incrementable<I> &&
	Sentinel<I, I>;
\end{codeblock}

\pnum
The domain of \tcode{==} for forward iterators is that of iterators over the same
underlying sequence. However, value-initialized iterators of the same type
may be compared and shall compare equal to other value-initialized iterators of the same type.
\begin{note}
	Value-initialized iterators behave as if they refer past the end of the same
	empty sequence.
\end{note}

\pnum
Pointers and references obtained from a forward iterator into a range \range{i}{s}
shall remain valid while \range{i}{s} continues to denote a range.

\pnum
Two dereferenceable iterators \tcode{a} and \tcode{b} of type \tcode{X}
offer the \defn{multi-pass guarantee} if:

\begin{itemize}
	\item \tcode{a == b} implies \tcode{++a == ++b} and
	\item The expression
	\tcode{((void)[](X x)\{++x;\}(a), *a)} is equivalent to the expression \tcode{*a}.
\end{itemize}

\pnum
\begin{note}
	The requirement that
	\tcode{a == b}
	implies
	\tcode{++a == ++b}
	and the removal of the restrictions on the number of assignments through
	a mutable iterator
	(which applies to output iterators)
	allow the use of multi-pass one-directional algorithms with forward iterators.
\end{note}


\rSec1[iterator.primitives]{Iterator primitives}

\rSec2[iterator.operations]{Iterator operations}

\pnum
Since only random access iterators provide
\tcode{+}
and
\tcode{-}
operators, the library provides two
function templates
\tcode{advance}
and
\tcode{distance}.
These
function templates
use
\tcode{+}
and
\tcode{-}
for random access iterators (and are, therefore, constant
time for them); for input, forward and bidirectional iterators they use
\tcode{++}
to provide linear time
implementations.

\indexlibrary{\idxcode{advance}}%
\begin{itemdecl}
	template<class InputIterator, class Distance>
	constexpr void advance(InputIterator& i, Distance n);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects
	\tcode{n}
	is negative only for bidirectional iterators.
	
	\pnum
	\effects
	Increments \tcode{i} by \tcode{n} if \tcode{n} is non-negative, and
	decrements \tcode{i} by \tcode{-n} otherwise.
\end{itemdescr}

\indexlibrary{\idxcode{distance}}%
\begin{itemdecl}
	template<class InputIterator>
	constexpr typename iterator_traits<InputIterator>::difference_type
	distance(InputIterator first, InputIterator last);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects
	\tcode{last} is reachable from \tcode{first}, or
	\tcode{InputIterator} meets
	the \oldconcept{RandomAccessIterator} requirements and
	\tcode{first} is reachable from \tcode{last}.
	
	\pnum
	\effects
	If \tcode{InputIterator} meets the \oldconcept{RandomAccessIterator} requirements,
	returns \tcode{(last - first)}; otherwise, returns
	the number of increments needed to get from
	\tcode{first}
	to
	\tcode{last}.
\end{itemdescr}

\indexlibrary{\idxcode{next}}%
\begin{itemdecl}
	template<class InputIterator>
	constexpr InputIterator next(InputIterator x,
	typename iterator_traits<InputIterator>::difference_type n = 1);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{advance(x, n); return x;}
\end{itemdescr}

\indexlibrary{\idxcode{prev}}%
\begin{itemdecl}
	template<class BidirectionalIterator>
	constexpr BidirectionalIterator prev(BidirectionalIterator x,
	typename iterator_traits<BidirectionalIterator>::difference_type n = 1);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{advance(x, -n); return x;}
\end{itemdescr}

\rSec2[range.iter.ops]{Range iterator operations}


\rSec3[range.iter.op.next]{\tcode{ranges::next}}

\indexlibrary{\idxcode{next}}%
\begin{itemdecl}
	template<@\added{Forward}@Iterator I>
	constexpr I ranges::next(I x);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{++x; return x;}
\end{itemdescr}

\indexlibrary{\idxcode{next}}%
\begin{itemdecl}
	template<@\added{Forward}Iterator I>
	constexpr I ranges::next(I x, iter_difference_t<I> n);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{ranges::advance(x, n); return x;}
\end{itemdescr}

\indexlibrary{\idxcode{next}}%
\begin{itemdecl}
	template<@\added{Forward}Iterator I, Sentinel<I> S>
	constexpr I ranges::next(I x, S bound);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{ranges::advance(x, bound); return x;}
\end{itemdescr}

\indexlibrary{\idxcode{next}}%
\begin{itemdecl}
	template<@\added{Forward}Iterator I, Sentinel<I> S>
	constexpr I ranges::next(I x, iter_difference_t<I> n, S bound);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{ranges::advance(x, n, bound); return x;}
\end{itemdescr}


\rSec2[move.iterators]{Move iterators and sentinels}

\pnum
Class template \tcode{move_iterator} is an iterator adaptor
with the same behavior as the underlying iterator except that its
indirection operator implicitly converts the value returned by the
underlying iterator's indirection operator to an rvalue.
Some generic algorithms can be called with move iterators to replace
copying with moving.

\pnum
\begin{example}
	
	\begin{codeblock}
		list<string> s;
		// populate the list \tcode{s}
		vector<string> v1(s.begin(), s.end());          // copies strings into \tcode{v1}
		vector<string> v2(make_move_iterator(s.begin()),
		make_move_iterator(s.end())); // moves strings into \tcode{v2}
	\end{codeblock}
	
\end{example}

\rSec3[move.iterator]{Class template \tcode{move_iterator}}

\indexlibrary{\idxcode{move_iterator}}%
\begin{codeblock}
	namespace std {
		template<class Iterator>
		class move_iterator {
			public:
			using iterator_type     = Iterator;
			using iterator_concept  = input_iterator_tag;
			using iterator_category = @\seebelow@;
			using value_type        = iter_value_t<Iterator>;
			using difference_type   = iter_difference_t<Iterator>;
			using pointer           = Iterator;
			using reference         = iter_rvalue_reference_t<Iterator>;
			
			constexpr move_iterator();
			constexpr explicit move_iterator(Iterator i);
			template<class U> constexpr move_iterator(const move_iterator<U>& u);
			template<class U> constexpr move_iterator& operator=(const move_iterator<U>& u);
			
			constexpr iterator_type base() const;
			@\added{constexpr iterator_type base() \&\&;}@
			constexpr reference operator*() const;
			
			constexpr move_iterator& operator++();
			constexpr auto operator++(int);
			constexpr move_iterator& operator--();
			constexpr move_iterator operator--(int);
			
			constexpr move_iterator operator+(difference_type n) const;
			constexpr move_iterator& operator+=(difference_type n);
			constexpr move_iterator operator-(difference_type n) const;
			constexpr move_iterator& operator-=(difference_type n);
			constexpr reference operator[](difference_type n) const;
			
			template<Sentinel<Iterator> S>
			friend constexpr bool
			operator==(const move_iterator& x, const move_sentinel<S>& y);
			template<Sentinel<Iterator> S>
			friend constexpr bool
			operator==(const move_sentinel<S>& x, const move_iterator& y);
			template<Sentinel<Iterator> S>
			friend constexpr bool
			operator!=(const move_iterator& x, const move_sentinel<S>& y);
			template<Sentinel<Iterator> S>
			friend constexpr bool
			operator!=(const move_sentinel<S>& x, const move_iterator& y);
			template<SizedSentinel<Iterator> S>
			friend constexpr iter_difference_t<Iterator>
			operator-(const move_sentinel<S>& x, const move_iterator& y);
			template<SizedSentinel<Iterator> S>
			friend constexpr iter_difference_t<Iterator>
			operator-(const move_iterator& x, const move_sentinel<S>& y);
			friend constexpr iter_rvalue_reference_t<Iterator>
			iter_move(const move_iterator& i)
			noexcept(noexcept(ranges::iter_move(i.current)));
			template<IndirectlySwappable<Iterator> Iterator2>
			friend constexpr void
			iter_swap(const move_iterator& x, const move_iterator<Iterator2>& y)
			noexcept(noexcept(ranges::iter_swap(x.current, y.current)));
			
			private:
			Iterator current;   // \expos
		};
	}
\end{codeblock}

\pnum
The member \grammarterm{typedef-name} \tcode{iterator_category} denotes
\begin{itemize}
	\item
	\tcode{random_access_iterator_tag} if
	the type
	\tcode{iterator_traits<\brk{}Iterator>::iterator_category} models
	\libconcept{DerivedFrom<\tcode{random_access_iterator_tag}>}, and
	\item
	\tcode{iterator_traits<\brk{}Iterator>::iterator_category} otherwise.
\end{itemize}

\rSec3[move.iter.requirements]{Requirements}

\pnum
The template parameter \tcode{Iterator} shall either
meet the \oldconcept{InputIterator} requirements\iref{input.iterators}
or model \libconcept{InputIterator}\iref{iterator.concept.input}.
Additionally, if any of the bidirectional traversal
functions are instantiated, the template parameter shall either
meet the \oldconcept{BidirectionalIterator} requirements\iref{bidirectional.iterators}
or model \libconcept{BidirectionalIterator}\iref{iterator.concept.bidir}.
If any of the random access traversal functions are instantiated, the
template parameter shall either
meet the \oldconcept{RandomAccessIterator} requirements\iref{random.access.iterators}
or model
\libconcept{RandomAccess\-Iterator}\iref{iterator.concept.random.access}.

\rSec3[move.iter.cons]{Construction and assignment}

\indexlibrary{\idxcode{move_iterator}!constructor}%
\begin{itemdecl}
	constexpr move_iterator();
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Constructs a \tcode{move_iterator}, value-initializing
	\tcode{current}. Iterator operations applied to the resulting
	iterator have defined behavior if and only if the corresponding operations are defined
	on a value-initialized iterator of type \tcode{Iterator}.
\end{itemdescr}


\indexlibrary{\idxcode{move_iterator}!constructor}%
\begin{itemdecl}
	constexpr explicit move_iterator(Iterator i);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Constructs a \tcode{move_iterator}, initializing
	\tcode{current} with \tcode{i}.
\end{itemdescr}


\indexlibrary{\idxcode{move_iterator}!constructor}%
\begin{itemdecl}
	template<class U> constexpr move_iterator(const move_iterator<U>& u);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\mandates \tcode{U} is convertible to \tcode{Iterator} \added{and \tcode{Iterator} is \tcode{CopyConstructible}}.
	
	\pnum
	\effects Constructs a \tcode{move_iterator}, initializing
	\tcode{current} with \tcode{u.base()}.
\end{itemdescr}

\indexlibrarymember{operator=}{move_iterator}%
\begin{itemdecl}
	template<class U> constexpr move_iterator& operator=(const move_iterator<U>& u);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\mandates \tcode{U} is convertible to \tcode{Iterator}.
	
	\pnum
	\effects Assigns \tcode{u.base()} to
	\tcode{current}.
\end{itemdescr}

\rSec3[move.iter.op.conv]{Conversion}

\indexlibrarymember{base}{move_iterator}%
\begin{itemdecl}
	constexpr Iterator base() const;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\begin{addedm}
		\constraints \tcode{CopyConstructible<Iterator>}
	\end{addedm}
	
	\returns \tcode{current}.
\end{itemdescr}



\begin{addedm}
\indexlibrarymember{base}{move_iterator}%
\begin{itemdecl}
	constexpr Iterator base() &&;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\returns \tcode{std::move(current);}
\end{itemdescr}
\end{addedm}

\rSec3[move.iter.elem]{Element access}

\indexlibrarymember{operator*}{move_iterator}%
\begin{itemdecl}
	constexpr reference operator*() const;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return ranges::iter_move(current);}
\end{itemdescr}

\indexlibrarymember{operator[]}{move_iterator}%
\begin{itemdecl}
	constexpr reference operator[](difference_type n) const;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{ranges::iter_move(current + n);}
\end{itemdescr}

\rSec3[move.iter.nav]{Navigation}

\indexlibrarymember{operator++}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator& operator++();
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects As if by \tcode{++current}.
	
	\pnum
	\returns \tcode{*this}.
\end{itemdescr}

\indexlibrarymember{operator++}{move_iterator}%
\begin{itemdecl}
	constexpr auto operator++(int);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\begin{addedm}
		\mandates
		\tcode{CopyConstructible<Iterator>}
	\end{addedm}

	\effects
	If \tcode{Iterator} models \libconcept{ForwardIterator}, equivalent to:
	\begin{codeblock}
		move_iterator tmp = *this;
		++current;
		return tmp;
	\end{codeblock}
	Otherwise, equivalent to \tcode{++current}.
\end{itemdescr}

\indexlibrarymember{operator\dcr}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator& operator--();
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects As if by \tcode{\dcr current}.
	
	\pnum
	\returns \tcode{*this}.
\end{itemdescr}

\indexlibrarymember{operator\dcr}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator operator--(int);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects
	As if by:
	\begin{codeblock}
		move_iterator tmp = *this;
		--current;
		return tmp;
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator+}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator operator+(difference_type n) const;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\returns \tcode{move_iterator(current + n)}.
\end{itemdescr}

\indexlibrarymember{operator+=}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator& operator+=(difference_type n);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects As if by: \tcode{current += n;}
	
	\pnum
	\returns \tcode{*this}.
\end{itemdescr}

\indexlibrarymember{operator-}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator operator-(difference_type n) const;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\returns \tcode{move_iterator(current - n)}.
\end{itemdescr}

\indexlibrarymember{operator-=}{move_iterator}%
\begin{itemdecl}
	constexpr move_iterator& operator-=(difference_type n);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects As if by: \tcode{current -= n;}
	
	\pnum
	\returns \tcode{*this}.
\end{itemdescr}

\rSec2[iterators.common]{Common iterators}

\rSec3[common.iterator]{Class template \tcode{common_iterator}}

\pnum
Class template \tcode{common_iterator} is an iterator/sentinel adaptor that is
capable of representing a non-common range of elements (where the types of the
iterator and sentinel differ) as a common range (where they are the same). It
does this by holding either an iterator or a sentinel, and implementing the
equality comparison operators appropriately.

\pnum
\begin{note}
	The \tcode{common_iterator} type is useful for interfacing with legacy
	code that expects the begin and end of a range to have the same type.
\end{note}

\pnum
\begin{example}
	\begin{codeblock}
		template<class ForwardIterator>
		void fun(ForwardIterator begin, ForwardIterator end);
		
		list<int> s;
		// populate the list \tcode{s}
		using CI = common_iterator<counted_iterator<list<int>::iterator>, default_sentinel_t>;
		// call \tcode{fun} on a range of 10 ints
		fun(CI(counted_iterator(s.begin(), 10)), CI(default_sentinel));
	\end{codeblock}
\end{example}

\indexlibrary{\idxcode{common_iterator}}%
\begin{codeblock}
	namespace std {
		template<Iterator I, Sentinel<I> S>
		requires (!Same<I, S>)
		class common_iterator {
			public:
			constexpr common_iterator() = default;
			constexpr common_iterator(I i);
			constexpr common_iterator(S s);
			template<class I2, class S2>
			requires ConvertibleTo<const I2&, I> && ConvertibleTo<const S2&, S>
			@\added{\&\& CopyConstructible<I>}@
			constexpr common_iterator(const common_iterator<I2, S2>& x);
			
			template<class I2, class S2>
			requires ConvertibleTo<const I2&, I> && ConvertibleTo<const S2&, S> &&
			Assignable<I&, const I2&> && Assignable<S&, const S2&>
			common_iterator& operator=(const common_iterator<I2, S2>& x);
			
			decltype(auto) operator*();
			decltype(auto) operator*() const
			requires @\placeholder{dereferenceable}@<const I>;
			decltype(auto) operator->() const
			requires @\seebelow@;
			
			common_iterator& operator++();
			decltype(auto) operator++(int);
			
			template<class I2, Sentinel<I> S2>
			requires Sentinel<S, I2>
			friend bool operator==(
			const common_iterator& x, const common_iterator<I2, S2>& y);
			template<class I2, Sentinel<I> S2>
			requires Sentinel<S, I2> && EqualityComparableWith<I, I2>
			friend bool operator==(
			const common_iterator& x, const common_iterator<I2, S2>& y);
			template<class I2, Sentinel<I> S2>
			requires Sentinel<S, I2>
			friend bool operator!=(
			const common_iterator& x, const common_iterator<I2, S2>& y);
			
			template<SizedSentinel<I> I2, SizedSentinel<I> S2>
			requires SizedSentinel<S, I2>
			friend iter_difference_t<I2> operator-(
			const common_iterator& x, const common_iterator<I2, S2>& y);
			
			friend iter_rvalue_reference_t<I> iter_move(const common_iterator& i)
			noexcept(noexcept(ranges::iter_move(declval<const I&>())))
			requires InputIterator<I>;
			template<IndirectlySwappable<I> I2, class S2>
			friend void iter_swap(const common_iterator& x, const common_iterator<I2, S2>& y)
			noexcept(noexcept(ranges::iter_swap(declval<const I&>(), declval<const I2&>())));
			
			private:
			variant<I, S> v_;   // \expos
		};
		
		template<class I, class S>
		struct incrementable_traits<common_iterator<I, S>> {
			using difference_type = iter_difference_t<I>;
		};
		
		template<InputIterator I, class S>
		struct iterator_traits<common_iterator<I, S>> {
			using iterator_concept = @\seebelow@;
			using iterator_category = @\seebelow@;
			using value_type = iter_value_t<I>;
			using difference_type = iter_difference_t<I>;
			using pointer = @\seebelow@;
			using reference = iter_reference_t<I>;
		};
	}
\end{codeblock}

\rSec3[common.iter.types]{Associated types}

\pnum
The nested \grammarterm{typedef-name}s of the specialization of
\tcode{iterator_traits} for \tcode{common_iterator<I, S>} are defined as follows.
\begin{itemize}
	\item
	\tcode{iterator_concept} denotes \tcode{forward_iterator_tag}
	if \tcode{I} models \libconcept{ForwardIterator};
	otherwise it denotes \tcode{input_iterator_tag}.
	
	\item
	\tcode{iterator_category} denotes
	\tcode{forward_iterator_tag}
	if \tcode{iterator_traits<I>::iterator_category}
	models \tcode{DerivedFrom<forward_iterator_tag>};
	otherwise it denotes \tcode{input_iterator_tag}.
	
	\item
	If the expression \tcode{a.operator->()} is well-formed,
	where \tcode{a} is an lvalue of type \tcode{const common_iterator<I, S>},
	then \tcode{pointer} denotes the type of that expression.
	Otherwise, \tcode{pointer} denotes \tcode{void}.
\end{itemize}

\rSec3[common.iter.const]{Constructors and conversions}

\indexlibrary{\idxcode{common_iterator}!constructor}%
\begin{itemdecl}
	constexpr common_iterator(I i);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects
	Initializes \tcode{v_} as if by \tcode{v_\{in_place_type<I>, std::move(i)\}}.
\end{itemdescr}

\indexlibrary{\idxcode{common_iterator}!constructor}%
\begin{itemdecl}
	constexpr common_iterator(S s);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Initializes \tcode{v_} as if by
	\tcode{v_\{in_place_type<S>, std::move(s)\}}.
\end{itemdescr}

\indexlibrary{\idxcode{common_iterator}!constructor}%
\begin{itemdecl}
	template<class I2, class S2>
	requires ConvertibleTo<const I2&, I> && ConvertibleTo<const S2&, S>
	@\added{\&\& CopyConstructible<I>}@
	constexpr common_iterator(const common_iterator<I2, S2>& x);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{x.v_.valueless_by_exception()} is \tcode{false}.
	
	\pnum
	\effects
	Initializes \tcode{v_} as if by
	\tcode{v_\{in_place_index<$i$>, get<$i$>(x.v_)\}},
	where $i$ is \tcode{x.v_.index()}.
\end{itemdescr}

\indexlibrarymember{operator=}{common_iterator}%
\begin{itemdecl}
	template<class I2, class S2>
	requires ConvertibleTo<const I2&, I> && ConvertibleTo<const S2&, S> &&
	Assignable<I&, const I2&> && Assignable<S&, const S2&>
	common_iterator& operator=(const common_iterator<I2, S2>& x);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{x.v_.valueless_by_exception()} is \tcode{false}.
	
	\pnum
	\effects
	Equivalent to:
	\begin{itemize}
		\item If \tcode{v_.index() == x.v_.index()}, then
		\tcode{get<$i$>(v_) = get<$i$>(x.v_)}.
		
		\item Otherwise, \tcode{v_.emplace<$i$>(get<$i$>(x.v_))}.
	\end{itemize}
	where $i$ is \tcode{x.v_.index()}.
	
	\pnum
	\returns \tcode{*this}
\end{itemdescr}

\rSec3[common.iter.access]{Accessors}

\indexlibrarymember{operator*}{common_iterator}%
\begin{itemdecl}
	decltype(auto) operator*();
	decltype(auto) operator*() const
	requires @\placeholder{dereferenceable}@<const I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{holds_alternative<I>(v_)}.
	
	\pnum
	\effects Equivalent to: \tcode{return *get<I>(v_);}
\end{itemdescr}

\indexlibrarymember{operator->}{common_iterator}%
\begin{itemdecl}
	decltype(auto) operator->() const
	requires @\seebelow@;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	The expression in the requires clause is equivalent to:
	\begin{codeblock}
		Readable<const I> &&
		(requires(const I& i) { i.operator->(); } ||
		is_reference_v<iter_reference_t<I>> ||
		Constructible<iter_value_t<I>, iter_reference_t<I>>)
	\end{codeblock}
	
	\pnum
	\expects \tcode{holds_alternative<I>(v_)}.
	
	\pnum
	\effects
	\begin{itemize}
		\item
		If \tcode{I} is a pointer type or if the expression
		\tcode{get<I>(v_).operator->()} is
		well-formed, equivalent to: \tcode{return get<I>(v_);}
		
		\item
		Otherwise, if \tcode{iter_reference_t<I>} is a reference type, equivalent to:
		\begin{codeblock}
			auto&& tmp = *get<I>(v_);
			return addressof(tmp);
		\end{codeblock}
		
		\item
		Otherwise, equivalent to:
		\tcode{return \placeholder{proxy}(*get<I>(v_));} where
		\tcode{\placeholder{proxy}} is the exposition-only class:
		\begin{codeblock}
			class @\placeholder{proxy}@ {
				iter_value_t<I> keep_;
				@\placeholder{proxy}@(iter_reference_t<I>&& x)
				: keep_(std::move(x)) {}
				public:
				const iter_value_t<I>* operator->() const {
					return addressof(keep_);
				}
			};
		\end{codeblock}
	\end{itemize}
\end{itemdescr}

\rSec3[common.iter.nav]{Navigation}

\indexlibrarymember{operator++}{common_iterator}%
\begin{itemdecl}
	common_iterator& operator++();
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{holds_alternative<I>(v_)}.
	
	\pnum
	\effects Equivalent to \tcode{++get<I>(v_)}.
	
	\pnum
	\returns \tcode{*this}.
\end{itemdescr}

\indexlibrarymember{operator++}{common_iterator}%
\begin{itemdecl}
	decltype(auto) operator++(int);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\begin{addedm}
		\mandates
		\tcode{CopyConstructible<I>}
	\end{addedm}

	\expects \tcode{holds_alternative<I>(v_)}.
	
	\pnum
	\effects
	If \tcode{I} models \libconcept{ForwardIterator}, equivalent to:
	\begin{codeblock}
		common_iterator tmp = *this;
		++*this;
		return tmp;
	\end{codeblock}
	Otherwise, equivalent to: \tcode{return get<I>(v_)++;}
\end{itemdescr}


\rSec2[iterators.counted]{Counted iterators}

\rSec3[counted.iterator]{Class template \tcode{counted_iterator}}

\pnum
Class template \tcode{counted_iterator} is an iterator adaptor
with the same behavior as the underlying iterator except that
it keeps track of the distance to the end of its range.
It can be used together with \tcode{default_sentinel}
in calls to generic algorithms to operate on
a range of $N$ elements starting at a given position
without needing to know the end position a priori.

\pnum
\begin{example}
	\begin{codeblock}
		list<string> s;
		// populate the list \tcode{s} with at least 10 strings
		vector<string> v;
		// copies 10 strings into \tcode{v}:
		ranges::copy(counted_iterator(s.begin(), 10), default_sentinel, back_inserter(v));
	\end{codeblock}
\end{example}

\pnum
Two values \tcode{i1} and \tcode{i2} of types
\tcode{counted_iterator<I1>}
and
\tcode{counted_iterator<I2>}
refer to elements of the same sequence if and only if
\tcode{next(i1.base(), i1.count())}
and
\tcode{next(i2.base(), i2.count())}
refer to the same (possibly past-the-end) element.

\indexlibrary{\idxcode{counted_iterator}}%
\begin{codeblock}
	namespace std {
		template<Iterator I>
		class counted_iterator {
			public:
			using iterator_type = I;
			
			constexpr counted_iterator() = default;
			constexpr counted_iterator(I x, iter_difference_t<I> n);
			template<class I2>
			requires ConvertibleTo<const I2&, I>
			constexpr counted_iterator(const counted_iterator<I2>& x);
			
			template<class I2>
			requires Assignable<I&, const I2&>
			constexpr counted_iterator& operator=(const counted_iterator<I2>& x);
			
			constexpr I base() const;
			@\added{constexpr I base() \&\&;}@
			constexpr iter_difference_t<I> count() const noexcept;
			constexpr decltype(auto) operator*();
			constexpr decltype(auto) operator*() const
			requires @\placeholder{dereferenceable}@<const I>;
			
			constexpr counted_iterator& operator++();
			decltype(auto) operator++(int);
			constexpr counted_iterator operator++(int)
			requires ForwardIterator<I>;
			constexpr counted_iterator& operator--()
			requires BidirectionalIterator<I>;
			constexpr counted_iterator operator--(int)
			requires BidirectionalIterator<I>;
			
			constexpr counted_iterator operator+(iter_difference_t<I> n) const
			requires RandomAccessIterator<I>;
			friend constexpr counted_iterator operator+(
			iter_difference_t<I> n, const counted_iterator& x)
			requires RandomAccessIterator<I>;
			constexpr counted_iterator& operator+=(iter_difference_t<I> n)
			requires RandomAccessIterator<I>;
			
			constexpr counted_iterator operator-(iter_difference_t<I> n) const
			requires RandomAccessIterator<I>;
			template<Common<I> I2>
			friend constexpr iter_difference_t<I2> operator-(
			const counted_iterator& x, const counted_iterator<I2>& y);
			friend constexpr iter_difference_t<I> operator-(
			const counted_iterator& x, default_sentinel_t);
			friend constexpr iter_difference_t<I> operator-(
			default_sentinel_t, const counted_iterator& y);
			constexpr counted_iterator& operator-=(iter_difference_t<I> n)
			requires RandomAccessIterator<I>;
			
			constexpr decltype(auto) operator[](iter_difference_t<I> n) const
			requires RandomAccessIterator<I>;
			
			template<Common<I> I2>
			friend constexpr bool operator==(
			const counted_iterator& x, const counted_iterator<I2>& y);
			friend constexpr bool operator==(
			const counted_iterator& x, default_sentinel_t);
			friend constexpr bool operator==(
			default_sentinel_t, const counted_iterator& x);
			
			template<Common<I> I2>
			friend constexpr bool operator!=(
			const counted_iterator& x, const counted_iterator<I2>& y);
			friend constexpr bool operator!=(
			const counted_iterator& x, default_sentinel_t y);
			friend constexpr bool operator!=(
			default_sentinel_t x, const counted_iterator& y);
			
			template<Common<I> I2>
			friend constexpr bool operator<(
			const counted_iterator& x, const counted_iterator<I2>& y);
			template<Common<I> I2>
			friend constexpr bool operator>(
			const counted_iterator& x, const counted_iterator<I2>& y);
			template<Common<I> I2>
			friend constexpr bool operator<=(
			const counted_iterator& x, const counted_iterator<I2>& y);
			template<Common<I> I2>
			friend constexpr bool operator>=(
			const counted_iterator& x, const counted_iterator<I2>& y);
			
			friend constexpr iter_rvalue_reference_t<I> iter_move(const counted_iterator& i)
			noexcept(noexcept(ranges::iter_move(i.current)))
			requires InputIterator<I>;
			template<IndirectlySwappable<I> I2>
			friend constexpr void iter_swap(const counted_iterator& x, const counted_iterator<I2>& y)
			noexcept(noexcept(ranges::iter_swap(x.current, y.current)));
			
			private:
			I current = I();                    // \expos
			iter_difference_t<I> length = 0;    // \expos
		};
		
		template<class I>
		struct incrementable_traits<counted_iterator<I>> {
			using difference_type = iter_difference_t<I>;
		};
		
		template<InputIterator I>
		struct iterator_traits<counted_iterator<I>> : iterator_traits<I> {
			using pointer = void;
		};
	}
\end{codeblock}

\rSec3[counted.iter.const]{Constructors and conversions}

\indexlibrary{\idxcode{counted_iterator}!constructor}%
\begin{itemdecl}
	constexpr counted_iterator(I i, iter_difference_t<I> n);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{n >= 0}.
	
	\pnum
	\effects
	Initializes \tcode{current} with \tcode{i} and
	\tcode{length} with \tcode{n}.
\end{itemdescr}

\indexlibrary{\idxcode{counted_iterator}!constructor}%
\begin{itemdecl}
	template<class I2>
	requires ConvertibleTo<const I2&, I>
	constexpr counted_iterator(const counted_iterator<I2>& x);
\end{itemdecl}

\begin{itemdescr}
	\pnum
\begin{addedm}
	\mandates
	\tcode{CopyConstructible<I>}
\end{addedm}
	
	\effects
	Initializes \tcode{current} with \tcode{x.current} and
	\tcode{length} with \tcode{x.length}.
\end{itemdescr}

\indexlibrarymember{operator=}{counted_iterator}%
\begin{itemdecl}
	template<class I2>
	requires Assignable<I&, const I2&>
	constexpr counted_iterator& operator=(const counted_iterator<I2>& x);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects
	Assigns \tcode{x.current} to \tcode{current} and
	\tcode{x.length} to \tcode{length}.
	
	\pnum
	\returns \tcode{*this}.
\end{itemdescr}

\rSec3[counted.iter.access]{Accessors}

\indexlibrarymember{base}{counted_iterator}%
\begin{itemdecl}
	constexpr I base() const;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\begin{addedm}
	\constraints \tcode{CopyConstructible<I>}
	\end{addedm}	
	
	\returns \tcode{current}.
\end{itemdescr}



\begin{addedm}
	\indexlibrarymember{base}{move_iterator}%
	\begin{itemdecl}
	
	constexpr I base() &&;
	\end{itemdecl}
	\begin{itemdescr}
		\pnum	
		\returns \tcode{std::move(current);}

	\end{itemdescr}

\end{addedm}


\indexlibrarymember{count}{counted_iterator}%
\begin{itemdecl}
	constexpr iter_difference_t<I> count() const noexcept;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return length;}
\end{itemdescr}

\rSec3[counted.iter.elem]{Element access}

\indexlibrarymember{operator*}{counted_iterator}%
\begin{itemdecl}
	constexpr decltype(auto) operator*();
	constexpr decltype(auto) operator*() const
	requires @\placeholder{dereferenceable}@<const I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return *current;}
\end{itemdescr}

\indexlibrarymember{operator[]}{counted_iterator}%
\begin{itemdecl}
	constexpr decltype(auto) operator[](iter_difference_t<I> n) const
	requires RandomAccessIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{n < length}.
	
	\pnum
	\effects Equivalent to: \tcode{return current[n];}
\end{itemdescr}

\rSec3[counted.iter.nav]{Navigation}

\indexlibrarymember{operator++}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator& operator++();
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{length > 0}.
	
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		++current;
		--length;
		return *this;
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator++}{counted_iterator}%
\begin{itemdecl}
	decltype(auto) operator++(int);
\end{itemdecl}

\begin{itemdescr}
	\pnum
\begin{addedm}
	\mandates \tcode{CopyConstructible<I>}
\end{addedm}
	
	\expects \tcode{length > 0}.
	
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		--length;
		try { return current++; }
		catch(...) { ++length; throw; }
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator++}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator operator++(int)
	requires ForwardIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		counted_iterator tmp = *this;
		++*this;
		return tmp;
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator{-}-}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator& operator--();
	requires BidirectionalIterator<I>
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		--current;
		++length;
		return *this;
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator{-}-}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator operator--(int)
	requires BidirectionalIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		counted_iterator tmp = *this;
		--*this;
		return tmp;
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator+}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator operator+(iter_difference_t<I> n) const
	requires RandomAccessIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return counted_iterator(current + n, length - n);}
\end{itemdescr}

\indexlibrarymember{operator+}{counted_iterator}%
\begin{itemdecl}
	friend constexpr counted_iterator operator+(
	iter_difference_t<I> n, const counted_iterator& x)
	requires RandomAccessIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return x + n;}
\end{itemdescr}

\indexlibrarymember{operator+=}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator& operator+=(iter_difference_t<I> n)
	requires RandomAccessIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{n <= length}.
	
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		current += n;
		length -= n;
		return *this;
	\end{codeblock}
\end{itemdescr}

\indexlibrarymember{operator-}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator operator-(iter_difference_t<I> n) const
	requires RandomAccessIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return counted_iterator(current - n, length + n);}
\end{itemdescr}

\indexlibrarymember{operator-}{counted_iterator}%
\begin{itemdecl}
	template<Common<I> I2>
	friend constexpr iter_difference_t<I2> operator-(
	const counted_iterator& x, const counted_iterator<I2>& y);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects
	\tcode{x} and \tcode{y} refer to elements of the same
	sequence\iref{counted.iterator}.
	
	\pnum
	\effects Equivalent to: \tcode{return y.length - x.length;}
\end{itemdescr}

\indexlibrarymember{operator-}{counted_iterator}%
\begin{itemdecl}
	friend constexpr iter_difference_t<I> operator-(
	const counted_iterator& x, default_sentinel_t);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to:
	\tcode{return -x.length;}
\end{itemdescr}

\indexlibrarymember{operator-}{counted_iterator}%
\begin{itemdecl}
	friend constexpr iter_difference_t<I> operator-(
	default_sentinel_t, const counted_iterator& y);
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\effects Equivalent to: \tcode{return y.length;}
\end{itemdescr}

\indexlibrarymember{operator-=}{counted_iterator}%
\begin{itemdecl}
	constexpr counted_iterator& operator-=(iter_difference_t<I> n)
	requires RandomAccessIterator<I>;
\end{itemdecl}

\begin{itemdescr}
	\pnum
	\expects \tcode{-n <= length}.
	
	\pnum
	\effects Equivalent to:
	\begin{codeblock}
		current -= n;
		length += n;
		return *this;
	\end{codeblock}
\end{itemdescr}

\section{Reference}
\begin{thebibliography}{9}
	
	\bibitem[P0541]{P0541}
	Eric Niebler
	\emph{ Ranges TS: Post-Increment on Input and Output Iterators}\newline
	\url{https://wg21.link/P0541}
	
	\bibitem[P0896]{P0896}
	Eric Niebler, Casey Carter, Christopher Di Bella.
	\emph{The One Ranges Proposal}\newline
	\url{https://wg21.link/P0896}
	
	\bibitem[P0902]{P0902}
	Andrew Hunter
	\emph{Move-only iterators}\newline
	\url{https://wg21.link/P0902}

	\bibitem[P1035]{P1035}
	Christopher Di Bella
	\emph{Input range adaptors}\newline
	\url{https://wg21.link/P1035}	
	
	
\end{thebibliography}

\end{document}