% !TeX program = luatex
% !TEX encoding = UTF-8

\RequirePackage{luatex85}%
\documentclass{wg21}

\usepackage{xcolor}
\usepackage{soul}
\usepackage{ulem}
\usepackage{fullpage}
\usepackage{parskip}
%\usepackage{listings}
\usepackage{enumitem}
\usepackage{makecell}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{amssymb}
\RequirePackageWithOptions{fontspec}
\usepackage{footnote}

\usepackage{newunicodechar}
\usepackage{url}
\usepackage[pass]{geometry}

\setmainfont{Noto Sans}



\usepackage{blindtext}
\usepackage[most]{tcolorbox}
\definecolor{block-gray}{gray}{0.95}


\newtcolorbox{quoteblock}[1][]{%
    colback=block-gray,
    grow to right by=-10mm,
    grow to left by=-10mm,
    boxrule=0pt,
    boxsep=0pt,
    breakable,
    enhanced jigsaw,
    borderline west={4pt}{0pt}{gray},
    #1,
}

\lstdefinestyle{MY}{language=C++,
    basicstyle=\small\CodeStyle,
    keywordstyle=\color{blue},
    commentstyle=\color{ForestGreen}\itshape\rmfamily,,
    stringstyle=\color{Sepia},
    xleftmargin=0em,
    showstringspaces=false,
    columns=flexible,
    keepspaces=true,
    keywords=[2]{char32_t, char8_t, char16_t},
    keywords=[3]{codepoint},
    keywords=[4]{noexcept,constexpr, static_assert},
    keywordstyle={\color{blue!80!black}},
    keywordstyle=[2]{\color{red!80!black}},
    keywordstyle=[3]{\color{BurntOrange!50!black}},
    keywordstyle=[4]{\color{green!50!black}},
    texcl=true}

\lstdefinestyle{MYSMALL}{language=C++,
    basicstyle=\footnotesize\CodeStyle,
    keywordstyle=\color{blue},
    commentstyle=\color{ForestGreen}\itshape,
    stringstyle=\color{Sepia},
    showstringspaces=false,
    columns=flexible,
    keepspaces=true,
    keywords=[2]{char32_t, char8_t, char16_t},
    keywords=[3]{codepoint},
    keywords=[4]{noexcept,constexpr, static_assert},
    keywordstyle={\color{blue!80!black}},
    keywordstyle=[2]{\color{red!80!black}},
    keywordstyle=[3]{\color{BurntOrange!50!black}},
    keywordstyle=[4]{\color{green!50!black}},
    texcl=true}


\newfontfamily{\fallbackfont}{Noto Sans}
\DeclareTextFontCommand{\textfallback}{\fallbackfont}
\newunicodechar{ẞ}{\textfallback{ẞ}}


% \setmonofont{Inconsolatazi4}
\setmainfont{Noto Serif}
\usepackage{csquotes}







\newcommand{\cc}[1]{\mintinline{c++}{#1}}


\title{Naming Text Encodings to Demystify Them}
\docnumber{D1885R2}
\audience{SG-16, LEWG}
\author{Corentin Jabot}{corentin.jabot@gmail.com}

\begin{document}
\maketitle

\begin{flushright}
    \hfill \break
    \hfill \break
    \textit{If you can't name it, you probably don't know what it is\\
    	If you don't know what it is, you don't know what it isn't\\
    	Tony Van Eerd}
\end{flushright}


\section{Target}

C++23

\section{Abstract}

For historical reasons, all text encodings mentioned in the standard are derived from
a locale object, which does not necessarily match the reality of how programs and system interact.

This model works poorly with modern understanding of text, ie the Unicode model separates encoding
from locales which are purely rules for formatting and text transformations but do not affect
which characters are represented by a sequence of code units.

Moreover, the standard does not provide a way to query which encodings are expected or used by the system,
leading to guesswork and unavoidable UB.

This paper introduces the notions of literal encoding, system encoding and a way to query them.


\section{Example}

\subsection{Listing the encoding}

\begin{lstlisting}[style=MY]
#include <text_encoding>
#include <iostream>

void print(const std::text_encoding & c) {
    std::cout << c.name()
    << " (iana mib: " << c.mib() << ")\n"
    << "Aliases:\n";
    for(auto && a : c.aliases()) {
        std::cout << '\t' << a << '\n';
    }
}

int main() {
    std::cout << "Literal Encoding: ";
    print(std::text_encoding::literal());
    std::cout << "Wide Literal Encoding: ";
    print(std::text_encoding::wide_literal());
    std::cout << "System Encoding: ";
    print(std::text_encoding::system());
    std::cout << "Wide system Encoding: ";
    print(std::text_encoding::wide_system());
}
\end{lstlisting}

Compiled with \tcode{g++ -fwide-exec-charset=EBCDIC-US -fexec-charset=SHIFT_JIS}, this program may display:

\begin{codeblock}
    Literal Encoding: SHIFT_JIS (iana mib: 17)
    Aliases:
        Shift_JIS
        MS_Kanji
        csShiftJIS
    
    Wide Literal Encoding: EBCDIC-US (iana mib: 2078)
    Aliases:
        EBCDIC-US
        csEBCDICUS
    
    System Encoding: UTF-8 (iana mib: 106)
    Aliases:
        UTF-8
        csUTF8
    
    Wide sytem Encoding: ISO-10646-UCS-4 (iana mib: 1001)
    Aliases:
        ISO-10646-UCS-4
        csUCS4
\end{codeblock}

\subsection{LWG3314}

[time.duration.io] specifies that the unit for micro seconds is µ on systems able to display it.
This is currently difficult to detect and implement properly.

The following allows an implementation to use µ if it is supported by both the execution encoding and the encoding
attached to the stream.


\begin{lstlisting}[style=MY]
    template<class traits, class Rep, class Period>
    void print_suffix(basic_ostream<char, traits>& os, const duration<Rep, Period>& d)
    {
        if constexpr(text_encoding::literal() == text_encoding::utf8) {
            if (os.getloc().encoding() == text_encoding::utf8) {
                os << d.count() << "\u00B5s"; // µ
                return;
            }
        }
        os << d.count() << "us";
    }
\end{lstlisting}

A more complex implementation may support more encodings, such as iso-8859-1.

\subsection{Asserting a specific encoding is set}

On POSIX, matching encodings is done by name, which pulls the entire database.
To avoid that we propose a method to asserting that the system encoding is as expected.
such method mixed to only pull in the strings associated with this encoding:

\begin{lstlisting}[style=MY]
int main() {
    return text_encoding::system_is<text_encoding::id::UTF8>();
}
\end{lstlisting}




\subsection{User construction}

To support other use cases such as interoperability with other libraries or internet protocols,
\tcode{text_encoding} can be constructed by users

\begin{lstlisting}[style=MY]
text_encoding my_utf8("utf8");
assert(my_utf8.name() == "utf8"sv); // Get the user provided name back
assert(my_utf8.mib() == text_encoding::id::UTF8);

text_encoding my_utf8_2(text_encoding::id::UTF8);
assert(my_utf8_2.name() == "UTF-8"sv); // Get the preferred name for the implementation
assert(my_utf8_2.mib() == text_encoding::id::UTF8);
assert(my_utf8 == my_utf8_2);
\end{lstlisting}

\subsection{Unregistered encoding}

Unregistered encoding are also supported. They have the \tcode{other} mib, no aliases and are compared
by names:

\begin{lstlisting}[style=MY]
text_encoding wtf8("WTF-8");
assert(wtf8.name() == "WTF-8"sv);
assert(wtf8.mib() == text_encoding::id::other);

//encodings with the \tcode{other} mib are compared by name, ignoring case, hyphens and underscores
assert(wtf8 == text_encoding("___wtf8__"));
\end{lstlisting}


\section{Revisions}

\subsection*{Revision 2}
\begin{itemize}
	\item Add all the enumerators of rcf 3008
    \item Add a mib constructor to \tcode{text_encoding}
    \item Add \tcode{system_is} and \tcode{system_wide_is} function templates
\end{itemize}

\subsection*{Revision 1}
\begin{itemize}
    \item Add more example and clarifications
    \item Require hosted implementations to support all the names registered in \cite{rfc3808}.
\end{itemize}

\section{Use cases}

This paper aims to make C++ simpler by exposing information that is currently hidden to the point of
being perceived as magical by many.
It also leaves no room for a language below C++ by ensuring that text encoding does not require the use of C functions.

The primary use cases are:

\begin{itemize}
    \item Ensuring a specific string encoding at compile time
    \item Ensuring at runtime that string literals are compatible with the system encoding
    \item Custom conversion function
    \item locale-independent text transformation
\end{itemize}

\subsection{Non goals}

This facility aims to help \textbf{identify} text encodings and does not want to solve encoding conversion and decoding.
Future text encoders and decoders may use the proposed facility as a way to identify their source and destination encoding.
The current facility is \emph{just} a fancy name.

\section{The many text encodings of a C++ system}

Text in a technical sense is a sequence of bytes to which is virtually attached an encoding.
Without encoding, a blob of data simply cannot be interpreted as text.

In many cases, the encoding used to encode a string is not communicated along with that string and its
encoding is therefore presumed with more or less success.

Generally, it is useful to know the encoding of a string when

\begin{itemize}
    \item Transferring data as text between systems or processes (I/O)
    \item Textual transformation of data
    \item Interpretation of a piece of data
\end{itemize}

In the purview of the standard, text I/O text originates from
\begin{itemize}
    \item The source code (literals)
    \item The iostream library as well as system functions
    \item Environment variables and command-line arguments intended to be interpreted as text.
\end{itemize}

Locales provide text transformation and conversion facilities and as such, in the current model have an encoding attached to them.

There are therefore 3 sets of encodings of primary interest:

\begin{itemize}
    \item The encoding of narrow and wide characters and string literals
    \item The narrow and wide encodings used by a program when sending or receiving strings from its environment
    \item The encoding of narrow and wide characters attached to a \tcode{std::locale} object
\end{itemize}

\note Because they have different code units sizes, narrow and wide strings have different encodings.
\tcode{char8_t}, \tcode{char16_t}, \tcode{char32_t} literals are assumed to be respectively UTF-8, UTF-16 and UTF-32 encoded.
\endnote

\note A program may have to deal with more encoding - for example, on Windows, the encoding of the console attached to \tcode{cout} may be different from the system encoding.

Likewise depending on the platform, paths may or may not have an encoding attached to them, and that encoding may either be a property of the platform or the filesystem itself.
\endnote


The standard only has the notion of execution character sets (which implies the existence of execution encodings), whose definitions are locale-specific.
That implies that the standard assumes that string literals are encoded in a subset of the encoding of the locale encoding.

This has to hold notably because it is not generally possible to differentiate runtime strings from compile-time literals at runtime.

This model does, however, present several shortcomings:

First, in practice, C++ software are often no longer compiled in the same environment as the one on which they are run and the entity providing the program may not have control over the environment on which it is run.

Both POSIX and C++ derives the encoding from the locale. Which is an unfortunate artifact of an era when 255 characters or less ought to be enough for anyone.
Sadly, the locale can change at runtime, which means the encoding which is used by ctype and conversion functions can change at runtime.
However, this encoding ought to be an immutable property as it is dictated by the environment (often the parent process).
In the general case, it is not for a program to change the encoding expected by its environment.
A C++ program sets the locale to "C" (see \cite{N2346}, 7.11.1.1.4) (which assumes a US ASCII encoding) during initialization, further losing information.


Many text transformations can be done in a locale-agnostic manner yet require the encoding to be known - as no text transformation can ever be applied without
prior knowledge of what the encoding of that text is.

More importantly, it is difficult or impossible for a developer to diagnose an incompatibility between the locale-derived, encoding, the system-assumed encoding and
the encoding of string literals.

Exposing the different encodings would let developers verify that that the system environment is compatible with the implementation-defined encoding of string literals, aka that the encoding and character set used to encode string literals are a strict subset of the encoding of the environment.

\section{Identifying Encodings}

To be able to expose the encoding to developers we need to be able to synthesize that information.
The challenge, of course, is that there exist many encodings (hundreds), and many names to refer to each one.
Fortunately there exist a database of registered encoding covering almost all encodings supported by operating systems and compilers.
This database is maintained by IANA through a process described by \cite{rfc2978}.

This database lists over 250 registered character sets and for each:
\begin{itemize}
    \item A name
    \item A unique identifier
    \item A set of known aliases
\end{itemize}

We propose to use that information to reliably identify encoding across implementations and systems.


\section{Design Considerations}

\subsection{Encodings are orthogonal to locales}

The following proposal is mostly independent of locales so that the relevant part can be implemented in an environment in which \tcode{<locale>}
is not available, as well as to make sure we can transition \tcode{std::locale} to be more compatible with Unicode.

\subsection{Naming}
SG-16 is looking at rewording the terminology associated with text and encoding throughout the standard, this paper does not yet reflect that effort.

However "system encoding" and "literal encoding" are descriptive terms. In particular "system" is illustrative of the
fact that a C++ program has, in the general case, no control over the encoding it is expected to produce and consume.

\subsection{MIBEnum}

We provide a \tcode{text_encoding::id} enum with the MIBEnum value of a few often used encodings for convenience.
Because there is a rather large number of encodings and because this list may evolve faster than the standard, it was pointed out during early review that it would be detrimental to attempt to provide a complete list.
\begin{note}
MIB stands for Management Information Base, which is IANA nomenclature, the name has no particular interest beside a desire not to deviate from the existing standards and practices.
\end{note}

The enum is purposefully not an \tcode{enum class} so that it can be easily compared to objects from
third party librairies such as \tcode{QTextCodec}.

The enumerators \tcode{unknown} and \tcode{other} and their value are provided by the very same RFC such as:

\begin{itemize}
	\item \tcode{other} designs an encoding not registered in the IANA Database, such that 2 encoding with the \tcode{other} mib are identical if their name compare equal.
	\item \tcode{unknown} is used when the encoding could not be determined. Under the current proposal,  only default constructing a \tcode{text_encoding} object can produce that value. The encoding associated with the
	locale or environment is always known.
\end{itemize}

While \tcode{MIBEnum} was necessary to make that proposal implementable consistently across platforms,
its main purpose is to remediate to the fact that an encoding can have multiple inconsistent names across implementations.

However,

\subsection{Name and aliases}

The proposed API offers both a name and aliases.
The \tcode{name} method reflects the name with which the \tcode{text_encoding} object was created, when applicable.
This is notably important when the encoding is not registered, or its name differs from the IANA name.

\subsection{Implementation flexibility}

This proposal aims to be implementable on all platforms as such, it supports encoding not registered with IANA, does not impose that
a freestanding implementation is aware of all registered encodings, and it let implementers provide their own aliases for IANA-registered encoding.
Because the process for registering encoding is documented \cite{rfc2978} implementations can (but are not required to) provide
registered encodings not defined in \cite{rfc3808} - in the case that rfc is updated out of sync of the standard.
However, \cite{rfc3808} is from 2004 and has not been updated. As the world converge to utf-8, new encodings are less likely to
be registered.

Implementations may not extend the \tcode{text_encoding::id} as to guarantee source compatibility.

\subsection{\tcode{const char*}}

A primary use case is to enable people to write their own conversion functions.
Unfortunately, most APIs expect NULL-terminated strings, which is why we return a const char*.

\section{Implementation}

The following proposal has been prototyped using a modified version of GCC to expose the encoding information.

On Windows, the run-time encoding can be determined by \tcode{GetACP} - and then map to MIB values, while on POSIX platform it corresponds to value of \tcode{nl_langinfo} when the environment ("") locale is set - before the program's locale is set to \tcode{C}.

On OSX \tcode{CFStringGetSystemEncoding} and \tcode{CFStringConvertEncodingToIANACharSetName} can also be used.


While exposing the literal encoding is novel, a few libraries do expose the system encoding, including Qt and wxWidget, and use the IANA registry.

Part of this proposal is available on \href{https://compiler-explorer.com/z/7D3Z3x}{Compiler explorer}
(literal and wide_literal are not supported)


\section{FAQ}
\subsection{Why not return a \tcode{text_encoding::id} rather than a \tcode{text_encoding} object?}
Some implementations may need to return a non-register encoding, in which case they would return \tcode{mib::other} and a custom name.

\tcode{text_encoding::system()} and \tcode{text_encoding::system_mib()} (not proposed) would generate the same code in an optimized build.


\subsection{But handling names is expensive?}
To ensure that the proposal is implementable in constrained environment, \tcode{text_encoding} has a limit of 63 characters per encoding name
which is sufficient to support all encodings we are aware of (registered or not)

\subsection{It seems like names and mib are separate concerns?}

Not all encodings are registered (event if most are), it is therefore not possible to identify uniquely all encoding uniquely by mib.
Encodings may have many names, but some platform will have a preferred name.

The combination of a name + a mib covers 100\% of use cases. Aliases further help with integration with third party libraries or to develop
tools that need mime encoding names.

\subsection{Why can't there be vendor provided MIBs?}

This would be meaningless in portable code. \tcode{mib} is only useful as mechanism to identify \textbf{portably} encoding and to increase compatibility across third party libraries.

It does not prevent the support of unregistered encodings:

\begin{codeblock}
    text_encoding wtf8("WTF-8");
    assert(wtf8.name() == "WTF-8"sv);
    assert(wtf8.mib() == text_encoding::id::other);
\end{codeblock}


\subsection{Why can't there be a \tcode{text_encoding(name, mib)} constructor?}

Same reason, if users are allowed to construct text_encoding from registered names or names otherwise
unknown from the implementation with an arbitrary mib, it becomes impossible to maintain the invariant of the class
(the relation between mib and name), which would make the interface much harder to use, without provided
any functionality.

\subsection{I just want to check that my platform is utf-8 without paying for all these other encoding?}

we added \tcode{system_is} to that end.

\begin{lstlisting}[style=MY]
    int main() {
        assert(text_encoding::system_is<text_encoding::id::UTF8>
                  && "Non UTF8 encoding detected, go away");
    }
\end{lstlisting}


This can be implemented in a way that only stores in the program the necessary information for that particular encoding (unless \tcode{aliases} is called at runtime).

On windows and OSX, only calling \tcode{encoding::aliases} would pull any data in the program, even if calling \tcode{system}.


\subsection{What is the cost of calling \tcode{aliases}?}

My crude implementation pulls in 30Ki of data when calling aliases or the name constructor, or \tcode{system()} (on POSIX).
\section{Future work}

Exposing the notion of text encoding in the core and library language gives us the tools to solve some
problems in the standard.

Notably, it offers a sensible way to do locale-independent, encoding-aware padding in \tcode{std::format} as in described in \cite{P1868}.

While this give us the tools to handle encoding, it does not fix the core wording.

\newpage


\section{Proposed wording}

Add the header \tcode{<text_encoding>} to the "C++ library headers" table in
\tcode{[headers]}, in a place that respects the table’s current alphabetic
order.

Add the macro \tcode{__cpp_lib_text_encoding} to \tcode{[version.syn]}, in a
place that respects the current alphabetic order:

\begin{codeblock}
#define __cpp_lib_text_encoding 201911L (**placeholder**) // also in \tcode{text_encoding}
\end{codeblock}

Add a new header \tcode{<text_encoding>}.

\begin{quote}
\begin{addedblock}

A \tcode{text\_encoding} describes a text encoding portably across platforms by exposing data from the Character Sets database described by \cite{rfc2978} and \cite{rfc3808}.

\begin{codeblock}

namespace std {

struct text_encoding final{
    enum class id : unsigned {
        other = 1,
        unknown = 2,
        ASCII = 3,
        ISOLatin1 = 4,
        ISOLatin2 = 5,
        ISOLatin3 = 6,
        ISOLatin4 = 7,
        ISOLatinCyrillic = 8,
        ISOLatinArabic = 9,
        ISOLatinGreek = 10,
        ISOLatinHebrew = 11,
        ISOLatin5 = 12,
        ISOLatin6 = 13,
        ISOTextComm = 14,
        HalfWidthKatakana = 15,
        JISEncoding = 16,
        ShiftJIS = 17,
        EUCPkdFmtJapanese = 18,
        EUCFixWidJapanese = 19,
        ISO4UnitedKingdom = 20,
        ISO11SwedishForNames = 21,
        ISO15Italian = 22,
        ISO17Spanish = 23,
        ISO21German = 24,
        ISO60DanishNorwegian = 25,
        ISO69French = 26,
        ISO10646UTF1 = 27,
        ISO646basic1983 = 28,
        INVARIANT = 29,
        ISO2IntlRefVersion = 30,
        NATSSEFI = 31,
        NATSSEFIADD = 32,
        NATSDANO = 33,
        NATSDANOADD = 34,
        ISO10Swedish = 35,
        KSC56011987 = 36,
        ISO2022KR = 37,
        EUCKR = 38,
        ISO2022JP = 39,
        ISO2022JP2 = 40,
        ISO13JISC6220jp = 41,
        ISO14JISC6220ro = 42,
        ISO16Portuguese = 43,
        ISO18Greek7Old = 44,
        ISO19LatinGreek = 45,
        ISO25French = 46,
        ISO27LatinGreek1 = 47,
        ISO5427Cyrillic = 48,
        ISO42JISC62261978 = 49,
        ISO47BSViewdata = 50,
        ISO49INIS = 51,
        ISO50INIS8 = 52,
        ISO51INISCyrillic = 53,
        ISO54271981 = 54,
        ISO5428Greek = 55,
        ISO57GB1988 = 56,
        ISO58GB231280 = 57,
        ISO61Norwegian2 = 58,
        ISO70VideotexSupp1 = 59,
        ISO84Portuguese2 = 60,
        ISO85Spanish2 = 61,
        ISO86Hungarian = 62,
        ISO87JISX0208 = 63,
        ISO88Greek7 = 64,
        ISO89ASMO449 = 65,
        ISO90 = 66,
        ISO91JISC62291984a = 67,
        ISO92JISC62991984b = 68,
        ISO93JIS62291984badd = 69,
        ISO94JIS62291984hand = 70,
        ISO95JIS62291984handadd = 71,
        ISO96JISC62291984kana = 72,
        ISO2033 = 73,
        ISO99NAPLPS = 74,
        ISO102T617bit = 75,
        ISO103T618bit = 76,
        ISO111ECMACyrillic = 77,
        ISO121Canadian1 = 78,
        ISO122Canadian2 = 79,
        ISO123CSAZ24341985gr = 80,
        ISO88596E = 81,
        ISO88596I = 82,
        ISO128T101G2 = 83,
        ISO88598E = 84,
        ISO88598I = 85,
        ISO139CSN369103 = 86,
        ISO141JUSIB1002 = 87,
        ISO143IECP271 = 88,
        ISO146Serbian = 89,
        ISO147Macedonian = 90,
        ISO150 = 91,
        ISO151Cuba = 92,
        ISO6937Add = 93,
        ISO153GOST1976874 = 94,
        ISO8859Supp = 95,
        ISO10367Box = 96,
        ISO158Lap = 97,
        ISO159JISX02121990 = 98,
        ISO646Danish = 99,
        USDK = 100,
        DKUS = 101,
        KSC5636 = 102,
        Unicode11UTF7 = 103,
        ISO2022CN = 104,
        ISO2022CNEXT = 105,
        UTF8 = 106,
        ISO885913 = 109,
        ISO885914 = 110,
        ISO885915 = 111,
        ISO885916 = 112,
        GBK = 113,
        GB18030 = 114,
        OSDEBCDICDF0415 = 115,
        OSDEBCDICDF03IRV = 116,
        OSDEBCDICDF041 = 117,
        ISO115481 = 118,
        KZ1048 = 119,
        UCS2 = 1000,
        UCS4 = 1001,
        UnicodeASCII = 1002,
        UnicodeLatin1 = 1003,
        UnicodeJapanese = 1004,
        UnicodeIBM1261 = 1005,
        UnicodeIBM1268 = 1006,
        UnicodeIBM1276 = 1007,
        UnicodeIBM1264 = 1008,
        UnicodeIBM1265 = 1009,
        Unicode11 = 1010,
        SCSU = 1011,
        UTF7 = 1012,
        UTF16BE = 1013,
        UTF16LE = 1014,
        UTF16 = 1015,
        CESU8 = 1016,
        UTF32 = 1017,
        UTF32BE = 1018,
        UTF32LE = 1019,
        BOCU1 = 1020,
        Windows30Latin1 = 2000,
        Windows31Latin1 = 2001,
        Windows31Latin2 = 2002,
        Windows31Latin5 = 2003,
        HPRoman8 = 2004,
        AdobeStandardEncoding = 2005,
        VenturaUS = 2006,
        VenturaInternational = 2007,
        DECMCS = 2008,
        PC850Multilingual = 2009,
        PC8DanishNorwegian = 2012,
        PC862LatinHebrew = 2013,
        PC8Turkish = 2014,
        IBMSymbols = 2015,
        IBMThai = 2016,
        HPLegal = 2017,
        HPPiFont = 2018,
        HPMath8 = 2019,
        HPPSMath = 2020,
        HPDesktop = 2021,
        VenturaMath = 2022,
        MicrosoftPublishing = 2023,
        Windows31J = 2024,
        GB2312 = 2025,
        Big5 = 2026,
        Macintosh = 2027,
        IBM037 = 2028,
        IBM038 = 2029,
        IBM273 = 2030,
        IBM274 = 2031,
        IBM275 = 2032,
        IBM277 = 2033,
        IBM278 = 2034,
        IBM280 = 2035,
        IBM281 = 2036,
        IBM284 = 2037,
        IBM285 = 2038,
        IBM290 = 2039,
        IBM297 = 2040,
        IBM420 = 2041,
        IBM423 = 2042,
        IBM424 = 2043,
        PC8CodePage437 = 2011,
        IBM500 = 2044,
        IBM851 = 2045,
        PCp852 = 2010,
        IBM855 = 2046,
        IBM857 = 2047,
        IBM860 = 2048,
        IBM861 = 2049,
        IBM863 = 2050,
        IBM864 = 2051,
        IBM865 = 2052,
        IBM868 = 2053,
        IBM869 = 2054,
        IBM870 = 2055,
        IBM871 = 2056,
        IBM880 = 2057,
        IBM891 = 2058,
        IBM903 = 2059,
        IBBM904 = 2060,
        IBM905 = 2061,
        IBM918 = 2062,
        IBM1026 = 2063,
        IBMEBCDICATDE = 2064,
        EBCDICATDEA = 2065,
        EBCDICCAFR = 2066,
        EBCDICDKNO = 2067,
        EBCDICDKNOA = 2068,
        EBCDICFISE = 2069,
        EBCDICFISEA = 2070,
        EBCDICFR = 2071,
        EBCDICIT = 2072,
        EBCDICPT = 2073,
        EBCDICES = 2074,
        EBCDICESA = 2075,
        EBCDICESS = 2076,
        EBCDICUK = 2077,
        EBCDICUS = 2078,
        Unknown8BiT = 2079,
        Mnemonic = 2080,
        Mnem = 2081,
        VISCII = 2082,
        VIQR = 2083,
        KOI8R = 2084,
        HZGB2312 = 2085,
        IBM866 = 2086,
        PC775Baltic = 2087,
        KOI8U = 2088,
        IBM00858 = 2089,
        IBM00924 = 2090,
        IBM01140 = 2091,
        IBM01141 = 2092,
        IBM01142 = 2093,
        IBM01143 = 2094,
        IBM01144 = 2095,
        IBM01145 = 2096,
        IBM01146 = 2097,
        IBM01147 = 2098,
        IBM01148 = 2099,
        IBM01149 = 2100,
        Big5HKSCS = 2101,
        IBM1047 = 2102,
        PTCP154 = 2103,
        Amiga1251 = 2104,
        KOI7switched = 2105,
        BRF = 2106,
        TSCII = 2107,
        CP51932 = 2108,
        windows874 = 2109,
        windows1250 = 2250,
        windows1251 = 2251,
        windows1252 = 2252,
        windows1253 = 2253,
        windows1254 = 2254,
        windows1255 = 2255,
        windows1256 = 2256,
        windows1257 = 2257,
        windows1258 = 2258,
        TIS620 = 2259,
        CP50220 = 2260,
        reserved = 3000
    };

    constexpr explicit text_encoding(string_view name);
    constexpr text_encoding(text_encoding::id mib) noexcept;

    constexpr id mib() const noexcept;
    constexpr const char* name() const noexcept;

    constexpr auto aliases() const noexcept -> @{\seebelow}@;

    constexpr bool operator==(const text_encoding & other) const;
    constexpr bool operator==(text_encoding::id mib) const;

    static consteval text_encoding literal();
    static consteval text_encoding wide_literal();

    static text_encoding system();
    static text_encoding wide_system();
    
    
    template<text_encoding::id id_>
    bool text_encoding::system_is();
    
    template<text_encoding::id id_>
    bool text_encoding::system_wide_is();

    private:
        id mib_; // \expos
        @\impldef{}@ name_; // \expos
    };


// hash support
template<class T> struct hash;
template<> struct hash<text_encoding>;

}


\end{codeblock}
\end{addedblock}
\end{quote}


\begin{addedblock}

\pnum A \defn{registered-character-set} is a character set registered by the process described in \cite{rfc2978} and which is known of the implementation.

\pnum Let \tcode{bool COMP_NAME(const char* a, const char* b)} be a function that returns \tcode{true} if two ASCII strings are identical, ignoring case and all \tcode{-} and  \tcode{_} characters.

\begin{note}
The enumerators of the \tcode{text_encoding::id} and their value match those specified in \cite{rfc3808} with the "cs" prefixed removed. \tcode{text_encoding::id::UCS2} corresponds to \tcode{csUnicode} in \cite{rfc3808}
\end{note}

\begin{itemdecl}
constexpr explicit text_encoding(string_view name);
\end{itemdecl}

\begin{itemdescr}
    \expects \tcode{name.size() < 64} is \tcode{true}.
    
    \effects
    If there exists an implementation-defined alias \tcode{a} of \defn{registered-character-set} such that \tcode{COMP_NAME(a, name.c_str())} is \tcode{true}, initialize
    \tcode{mib_} with the \tcode{MIBenum} associated with that \defn{registered-character-set}. Otherwise, initialize \tcode{mib_} with \tcode{text_encoding::id::other}.

    Implementations must return a valid \tcode{text_encoding} object for every \tcode{name} that matches either an alias or a name of a \defn{registered-character-set} listed in \cite{rfc3808}.

\begin{note}
	Freestanding implementations are not required to provide this method
\end{note}

    \ensures \tcode{name_ == name}.
\end{itemdescr}


\begin{itemdecl}
constexpr text_encoding(text_encoding::id mib) noexcept;
\end{itemdecl}

\begin{itemdescr}
    \expects \tcode{mib}  has the value of one of the enumerators of \tcode{text_encoding::id}.
    
    \ensures \tcode{mib_ == mib}.
\end{itemdescr}





\begin{itemdecl}
constexpr id mib() const noexcept;
\end{itemdecl}

\begin{itemdescr}
\returns \tcode{mib\_}.

\begin{note}
	The enumerator value \tcode{text_encoding::id::unknown} is provided for compatibility with \cite{rfc3808},
	\tcode{text_encoding::mib()} never returns \tcode{text_encoding::id::unknown}.
\end{note}

\end{itemdescr}

\begin{itemdecl}
constexpr const char* name() const noexcept;
\end{itemdecl}
\begin{itemdescr}
\returns 

\begin{itemize}
    \item \tcode{name_} if \tcode{strlen(name_) > 0} is \tcode{true},
    \item Otherwise, if \tcode{id != id::unknown} is \tcode{true}, an implementation defined null-terminated string corresponding to the preferred name of the encoding on that platform.
    \item Otherwise, \tcode{nullptr}
\end{itemize}
\end{itemdescr}


\begin{itemdecl}
constexpr auto aliases() const noexcept;
\end{itemdecl}

\begin{itemdescr}
\returns an implementation-defined object \tcode{r} representing a sequence of aliases such that:
\begin{itemize}
    \item \tcode{ranges::view<decltype(r)>} is true,
    \item \tcode{ranges::random_access_range<decltype(r)>} is true,
    \item \tcode{same_as<ranges::range_value_t<decltype(r)>, string_view>} is true,
    \item \tcode{!ranges::empty(r) || mib() == id::other} is true.
\end{itemize}

\pnum If \tcode{mib()} is equal to the \tcode{MIBEnum} value of one of the \defn{registered-character-sets}, \tcode{r[0]} is the name of the \defn{registered-character-set}.

\tcode{r} contains the aliases of the \defn{registered-character-set} as specified by \cite{rfc2978}.

\pnum \tcode{r} may contain implementation-defined values.

\pnum \tcode{r} does not contain duplicated values - the equality of 2 values is determined by \tcode{COMP_NAME}.

\begin{note}
The order of elements in \tcode{r} is unspecified.
\end{note}

\end{itemdescr}

\begin{itemdecl}
constexpr bool operator==(const text_encoding & other) const;
\end{itemdecl}

\begin{itemdescr}
\returns
\tcode{COMP_NAME(name(), other.name())} if \tcode{mib() == id::other \&\& other.mib() == id::other} is true,
otherwise
\tcode{mib() == other.mib()}.
\end{itemdescr}

\begin{itemdecl}
constexpr bool operator==(text_encoding::id i) const;
\end{itemdecl}

\begin{itemdescr}
\returns \tcode{(mib() != id::other) ? mib() == i : false}.
\end{itemdescr}

\begin{itemdecl}
static consteval text_encoding literal();
\end{itemdecl}

\begin{itemdescr}
    \returns a \tcode{text_encoding} object representing the narrow execution encoding.
\end{itemdescr}

\begin{itemdecl}
static consteval text_encoding wide_literal();
\end{itemdecl}

\begin{itemdescr}
    \returns a \tcode{text_encoding} object representing the wide execution encoding.
\end{itemdescr}

\begin{itemdecl}
static text_encoding system();
\end{itemdecl}

\begin{itemdescr}
Return the presumed system narrow encoding.
On POSIX systems this is the encoding attached to the environment locale (\tcode{""}) at the start of the program.

\begin{note}
This function should always return the same value during the lifetime of a program and is not affected by calls to \tcode{setlocale}.
\end{note}
\end{itemdescr}

\begin{itemdecl}
static text_encoding wide_system();
\end{itemdecl}

\begin{itemdescr}
Return the presumed system wide encoding.
On POSIX systems this is the encoding attached to the environment locale (\tcode{""}) at the start of the program.

\begin{note}
This function should always return the same value during the lifetime of a program and is not affected by calls to \tcode{setlocale}.
\end{note}
\end{itemdescr}

\begin{itemdecl}
template<text_encoding::id id_>
static bool text_encoding::system_is();
\end{itemdecl}

\begin{itemdescr}
\returns Equivalent to \tcode{system() == id_}
\end{itemdescr}

\begin{itemdecl}
template<text_encoding::id id_>
static bool text_encoding::system_wide_is();
\end{itemdecl}

\begin{itemdescr}
\returns Equivalent to \tcode{system_wide() == id_}
\end{itemdescr}


\end{addedblock}

In \tcode{[locale]}:

\begin{quote}
\begin{codeblock}
namespace std {
  class locale {
  public:
    [...]

    // locale operations
    string name() const;
\end{codeblock}
\begin{addedblock}
\begin{codeblock}
    text_encoding encoding() const;
    text_encoding wide_encoding() const;
\end{codeblock}
\end{addedblock}
\begin{codeblock}
  };
}
\end{codeblock}
\end{quote}

In \tcode{[locale.members]}:

\begin{itemdecl}
string name() const;
\end{itemdecl}

\begin{itemdescr}
\pnum
\returns
The name of
\tcode{*this},
if it has one; otherwise, the string \tcode{"*"}.
\end{itemdescr}

\begin{addedblock}
\begin{itemdecl}
text_encoding encoding() const;
\end{itemdecl}

\begin{itemdescr}
  \returns The text encoding for narrow strings associated with the locale
  \tcode{*this}.
\end{itemdescr}

\begin{itemdecl}
text_encoding wide_encoding() const;
\end{itemdecl}

\begin{itemdescr}
  \returns The text encoding for wide strings associated with the locale
  \tcode{*this}.
\end{itemdescr}
\end{addedblock}

\section{Acknowledgments}

Many thanks to Victor Zverovich, Thiago Macieira, Jens Maurer, Tom Honermann and others for reviewing this work and providing valuable feedback.


\begin{thebibliography}{9}

    \bibitem[N4830]{N4830}
    Richard Smith
    \emph{Working Draft, Standard for Programming Language C++}\newline
    \url{https://wg21.link/n4830}


    \bibitem[N2346]{N2346}
    \emph{Working Draft, Standard for Programming Language C}\newline
    \url{ http://www.open-std.org/jtc1/sc22/wg14/www/docs/n2346.pdf}




    \bibitem[rfc3808]{rfc3808}
    I. McDonald
    \emph{IANA Charset MIB}\newline
    \url{https://tools.ietf.org/html/rfc3808}

    \bibitem[rfc2978]{rfc2978}
    N. Freed
    \emph{IANA Charset MIB}\newline
    \url{https://tools.ietf.org/html/rfc2978}

    \bibitem[Character Sets]{mib}
    IANA
    \emph{Character Sets}\newline
    \url{https://www.iana.org/assignments/character-sets/character-sets.xhtml}


    \bibitem[iconv encodings]{iconv}
    GNU project
    \emph{Iconv Encodings}\newline
    \url{http://git.savannah.gnu.org/cgit/libiconv.git/tree/lib/encodings.def}

    \bibitem[P1868]{P1868}
    Victor Zverovich
    \emph{Clarifying units of width and precision in std::format}\newline
    \url{http://wg21.link/P1868}



\end{thebibliography}

\end{document}
