% !TeX program = luatex
% !TEX encoding = UTF-8


\documentclass{wg21}

\title{\tcode{constexpr} structured bindings\\{\small and}\\\ references to \tcode{constexpr} variables}
\docnumber{D2686R3}
\audience{EWG, CWG}
\author{Corentin Jabot}{corentin.jabot@gmail.com}
\authortwo{Brian Bi}{bbi10@bloomberg.net}

\usepackage{color, colortbl}
\begin{document}
\maketitle

\section{Abstract}

\paper{P1481R0} proposed allowing references to constant expressions to be themselves constant expressions,
as a means to support \tcode{constexpr} structured bindings.
This paper reports implementation experience on this proposal and provides updated wording.


\section{Revisions}

\subsection{Revision 3}

CWG pointed out that the restrictions on volatile variables and mutable subobjects in the definition of ``usable in constant expressions'' have undesirable consequences. Those restrictions have therefore been moved to [basic.def.odr]. An unintentional omission in R2 has also been fixed: a local static constexpr reference shall not refer to a variable with automatic storage duration. Examples have also been added to the wording section.

Following discussion in EWG, we have also expanded the rationale for the restriction that constexpr references can only refer to automatic storage duration variables that have the same innermost enclosing function.

\subsection{Revision 2}

We provide wording for option 3 (symbolic addressing), which is the direction chosen by EWG in Varna.
We also allow \tcode{constinit} structured bindings, mostly because we could not find a good reason not to,
and we think it's best to avoid too many exceptions and inconsistencies.

\subsection{Revision 1}

After core expressed implementability concerns of the original design as it pertains to constexpr references to automatic storage duration variables, we provide different options.

\subsection{Revision 0}

Design and wording similar to that of \paper{P1481R0}.

\section{Issues with R0 and possible solutions}

The previous revision of this paper, (\paper{P2686R0}), was approved by the EWG in Issaquah and was subsequently
reviewed by CWG, which found the proposed wording to be quite insufficient.

No issue arises with allowing \tcode{constexpr} structured binding in general, except for the case of an automatic storage duration structured binding
initialized by a tuple, i.e.,

\begin{colorblock}
void f() {
    constexpr auto [a] = std::tuple(1);
    static_assert(a == 1);
}
\end{colorblock}

which translates to

\begin{colorblock}
void f() {
    constexpr auto __sb = std::tuple(1);  // __sb has automatic storage scenario.
    constexpr const int& a = get<0>(__sb);
}
\end{colorblock}

When the structured binding is over an array or a class type, it doesn't create actual references,
so we have no issue. When the structured binding is not at function scope, the underlying tuple object has
static storage duration, and its address is a permitted result of a constant expression.

So the problematic case occurs when we are creating an automatic storage duration (i.e., at block scope) structured binding of a tuple (or \placeholder{tuple-like})
object. This specific situation, though, is not uncommon.

The initial wording simply allowed references initialized by a constant expression to be usable in constant expressions.
This phrasing failed to observe that the address of a \tcode{constexpr} variable with automatic storage duration may be different for each evaluation
of a function and, therefore, cannot be a \emph{permitted result of a constant expression}.

The CWG asks that the EWG consider and pick one direction to resolve these concerns.
Some options are explored below.

\section{Possible solutions}

\subsection{0. Allowing static and non-tuple \tcode{constexpr} structured binding}

We should be clear that nothing prevents \tcode{constexpr} structured bindings from just working when binding an aggregate or an array since those are modeled by special magic aliases that are not quite references
(which allows them to work with bitfields).

A \tcode{constexpr} structured binding of a tuple \emph{with static storage duration}, i.e.,

\begin{colorblock}
static constexpr auto [a, b] = std::tuple{1, 2};
\end{colorblock}

would also simply work as it would be equivalent to

\begin{colorblock}
static constexpr auto __t = std::tuple{1, 2};
static constexpr auto & a = std::get<0>(__t);
static constexpr auto & b = std::get<1>(__t);
\end{colorblock}

Supporting this solution requires no further changes to the language than basically allowing the compiler to
parse and apply the \tcode{constexpr} specifier.
Independently of the other solutions presented here, this option would be useful and should be done.

The problematic scenario is an automatic storage duration binding to a \tcode{tuple}.

We could stop there, not try to solve this problem, and force users to use \tcode{static}.
We would, however, have to ensure that expansion statements work with static variables since that was one of the motivations for this paper.

\subsection{1. Making \tcode{constexpr} implicitly static}

We could make \tcode{constexpr} variables implicitly static, but
doing so would most certainly break existing code, in addition to being inconsistent with the meaning of \tcode{constexpr}:

\begin{colorblock}
int f() {
    constexpr struct S {
        mutable int m ;
    } s{0};
    return ++s.m;
}

int main() {
    assert(f() + f() == 2); // currently 2. Becomes 3 if 's' is made implicitly static
}
\end{colorblock}

So this solution is impractical. We could make \tcode{constexpr} static only in some cases to alleviate some of the breakages or even make only \tcode{constexpr} bindings static, not other variables, but this option feels like a hack rather than an actual solution.

\subsection{2. Always re-evaluate a call to \tcode{get}?}

We could conceive that during constant evaluation, tuple structured bindings are replaced by a call to \tcode{get} every time they are constant-evaluated.
This would help with \tcode{constexpr} structured binding but would still disallow generic cases:

\begin{colorblock}
constexpr in not_a_sb =1;
constexpr const int&  a = sb;
\end{colorblock}

Additionally, this would be observable in scenarios in which \tcode{get} would perform some kind of compile-time i/o such as proposed by \paper{P2758R0}.

\subsection{3. Symbolic addressing}

The most promising option --- the one we think should be pursued --- is for \tcode{constexpr} references to designate a specific object, rather than an address,
and to retain that information across constant evaluation contexts.
This is how constant evaluation of references works, but this information is not currently persisted across constant evaluation, which is why we do not permit
\tcode{constexpr} references to refer to objects with automatic storage duration (or subobjects thereof).

To quote \href{https://lists.isocpp.org/core/2023/04/14163.php}{a discussion on the reflector}:

\begin{quoteblock}
This would also resolve a longstanding complaint that the following is invalid:
\begin{colorblock}
void f() {
    constexpr int a = 1;
    constexpr auto *p = &a;
}
\end{colorblock}

It seems like a lot of C++ developers expect the declaration of p to be valid, even though it's potentially initialized to a different address each time f is invoked.
\end{quoteblock}

This solution has the benefit of not being structured-binding specific and would arguably meet user expectations better than the current rule.
Interestingly and maybe counter-intuitively, the constexprness of pointers and references is completely orthogonal to that of their underlying object:

\begin{colorblock}
int main() {
    static int i = 0;
    static constexpr int & r = i; // currently valid

    int j = 0;
    constexpr int & s = j; // could be valid under the "symbolic addressing" model
}
\end{colorblock}

References can be constant expressions because we can track during constant evaluation which objects they refer to, independently of whether the value of that object is or isn't a constant expression.

We would have to be careful about several things.
Pointers and references to variables with automatic storage duration cannot be used outside of the lifetime of their underlying objects, so they could not appear
\begin{itemize}
  \item in template arguments
  \item as the initializer of a variable with static storage duration
\end{itemize}

Similarly, we can construct an automatic storage duration \tcode{constexpr} reference to a static variable but not a static \tcode{constexpr} reference bound to an automatic storage duration object.

\subsection{Thread-local variables}

Taking the address of a thread-local variable may initialize the variable, and that initialization may not be a constant expression.
Supporting references/pointers to thread-local variables would therefore require additional consideration, and we would probably want to allow it only if it were  already initialized
on declaration.

We could exclude thread locals from the design entirely as we're not sure a compelling use case exists for constexpr references to thread-local objects.

\subsection{Lambdas that reference automatic storage duration objects from the enclosing function}

\tcode{constexpr} references are not ODR-used.  Therefore, a constexpr reference used in a lambda does not trigger a capture.
This would be problematic for references bound to automatic storage duration objects:

\begin{colorblock}
auto f() {
    int i = 0;
    constexpr const int & ref = i;
    return [] {
        return ref;
    });
}
f();
\end{colorblock}

The current rules never require odr-use of \tcode{constexpr} references because
any use of a \tcode{constexpr} reference can always be evaluated as if it were
replaced by its initializer. However, in the above example, such a
transformation is not desirable since it would require \tcode{i} to be captured.
Instead, we need to
modify \href{http://eel.is/c++draft/basic.def.odr#5.1}{[basic.def.odr]/p5.1} so that \tcode{constexpr} references to automatic storage duration variables (or subobjects thereof) are ODR-used.
In the above example, \tcode{ref} therefore needs to be captured.

Lambdas also present other design questions. During EWG review in Kona (November
2023), examples similar to the following were discussed:
\begin{colorblock}
auto f1() {
    int i = 0;
    constexpr const int & ref = i;
    return [&] {
        constexpr const int & ref2 = i;
        return ref2;
    }();
}
auto f2() {
    int i = 0;
    constexpr const int & ref = i;
    return [&] {
        constexpr const int & ref2 = i;
        return ref2;
    };
}
\end{colorblock}
We propose that when a \tcode{constexpr} reference binds to a variable with
automatic storage duration, that variable must belong to the same innermost
function definition as the reference in order to ensure that \tcode{constexpr}
references cannot become dangling as a result of the referenced entity going out
of scope. Therefore, we propose that the definitions of \tcode{f1} and
\tcode{f2} both be ill formed.

Some participants in EWG expressed the sentiment that \tcode{f1} should be
well-formed but \tcode{f2} should be ill formed. In \tcode{f1}, the reference
\tcode{ref2} never bceomes dangling because the lambda is invoked immediately,
while in \tcode{f2}, the lambda may be invoked by the caller of \tcode{f2}, at
which point the behavior is necessarily undefined. Specifically, the concern was
raised that the restrictions that we propose are difficult to teach. We
therefore discuss and evaluate a number of suboptions of Option 3.

\subsubsection{Referenced variable must be in the same function (Suboption 3a)}
The suboption we propose (and provide wording for below) is that a
\tcode{constexpr} reference cannot bind to a variable with automatic storage
duration unless that variable is defined in the same innermost function. This
makes both \tcode{f1} and \tcode{f2} ill formed. The advantages of this approach
are:
\begin{itemize}
\item It is easy to specify.
\item In every case where a \tcode{constexpr} reference might become dangling
as a result of the referenced entity going out of scope, a diagnostic is
guaranteed.
\item The diagnostic is easy to understand.
\item It leaves open the possibility of the restrictions being relaxed by a
future proposal (i.e., if \tcode{f1} is not allowed now, it can be made allowed
later).
\end{itemize}
The main disadvantage of this approach is that it is conservative in terms of
what it accepts. In \tcode{f1}, the reference inside the lambda will never be
dangling, but this approach rejects it anyway. While it is easy to describe the
restriction to C++ learners, it is difficult to teach the rationale.

\subsubsection{Specified escape analysis (Suboption 3b)}
It may be desirable for \tcode{f1} to be well formed but \tcode{f2} ill formed
because in \tcode{f1}, the \tcode{constexpr} reference can never dangle (because
the lambda doesn't ``escape'' the function that \tcode{i} belongs to) whereas in
\tcode{f2} the reference can dangle (and, in fact, always does). But it is not
possible for a compiler to accept all \tcode{constexpr} references that never
dangle while rejecting all \tcode{constexpr} references that potentially dangle,
because doing so would require solving arbitrary decision problems. In other
words, suboption 3b only has the potential to get us \emph{closer} to the
``line'' between all \tcode{constexpr} variables being safe to use and all safe
uses of \tcode{constexpr} variables being accepted by the compiler; it's not
possible to land exactly \emph{on} that line.

Instead, the following specification strategy could be pursued. Define the term
\emph{constexpr-closes over}, where a lambda constexpr-closes over any variable
that is bound to any \tcode{constexpr} reference defined in the lambda body. So
in both \tcode{f1} and \tcode{f2}, the lambdas potentially constexpr-close over
the respective \tcode{i} variables. Then, specify a set of core language rules
that define when an expression or variable \emph{potentially references} a
lambda. For example, a ternary expression potentially references any lambda that
is potentially referenced by its second or third operand, and a reference
variable potentially references any lambda that is potentially referenced by its
initializer. Finally, we would say that a \tcode{return} statement is ill
formed if the returned expression potentially references any lambda that
constexpr-closes over any variable with automatic storage duration that is
defined in the function. In \tcode{f1} above, the rules would eventually
determine that the \tcode{return} statement inside the lambda doesn't
potentially reference the lambda itself, which implies that the result of
invoking the lambda doesn't potentially reference the lambda. Thus, as long as
only the result of invoking the lambda is returned from \tcode{f1}, the returned
expression from \tcode{f1} doesn't potentially reference any lambda that
constexpr-closes over \tcode{i}. However, in \tcode{f2}, the returned expression
does potentially reference the lambda, so its \tcode{return} statement would be
ill formed.

We would also need to introduce restrictions that would prevent copies of or
references to such lambdas from leaking out of the function in other ways, such
as by being assigned to a global variable with a \tcode{std::function} type.

The main advantages of this approach are:
\begin{itemize}
\item In every case where a \tcode{constexpr} reference might become dangling
as a result of the referenced entity going out of scope, a diagnostic is
guaranteed.
\item It allows a wider range of well-formed code than suboption 3a.
\end{itemize}
The main disadvantages are:
\begin{itemize}
\item It is extraordinarily difficult to specify, requiring careful analysis of
all constructs that potentially allow references to escape.
\item It burdens implementations by requiring them to track the sets of
potentially referenced lambdas.
\item It still has a teachability burden. This is because the specification will
necessarily result in some false positives, i.e., situations in which the human
programmer knows that the \tcode{constexpr} reference doesn't escape, but the
specification says that the containing lambda \emph{is} potentially referenced.
In those situations, the programmer will need to understand the path through
which the alleged potential reference escapes, and restructure their code in
order to satisfy the compiler.
\item The diagnostics that would be produced in both true positive and false
positive cases would not be easy to understand. If they just said that a
particular local variable potentially escapes, the programmer would need to
figure out why. On the other hand, the diagnostic could actually describe the
path, but such a diagnostic would not necessarily be easy to read. (It could be
compared with e.g. current diagnostics that attempt to explain why a constraint
on a templated entity is not satisfied or recursively references itself.)
\end{itemize}

\subsubsection{Allow \tcode{constexpr} references to dangle (Suboptions 3c and 3d)}
We could \emph{not} introduce the same-function restriction, and thus permit
both \tcode{f1} and \tcode{f2}. In this case, if the user ever invokes the
lambda returned from \tcode{f2} at runtime, the behavior is undefined. However,
the fact that the definition of \tcode{f2} can ever compile is strange, because
any invocation of the lambda returned from \tcode{f2} would result in control
flow passing through the initialization of \tcode{ref2}, and such an
initialization would fail to be a constant expression (because binding a
reference to an object that doesn't exist has undefined behavior). So in this
particular case, the fact that the initialization of a \tcode{constexpr}
variable is not a constant expression would not be diagnosed---nor could it be,
because that initialization is not evaluated until runtime!

In current C++, a \tcode{constexpr} reference \tcode{r} with automatic storage
duration is, in the abstract machine, initialized any number of times (depending
on how many times control flow passes through its definition) but the
implementation really evaluates it only once: i.e., it evaluates the initializer
of \tcode{r} as if \tcode{r} were a variable with static storage duration (and
thus subject to constant initialization); and whatever object \tcode{o} the
resulting reference ends up referring to, every \grammarterm{id-expression}
denoting \tcode{r} is transformed into one that denotes \tcode{o}. If we pursue
suboption 3c, then in order to handle the case where \tcode{r} may refer to an
object with automatic storage duration, we will need to actually change the
\emph{abstract machine rules} to say that whether or not the initialization of
\tcode{r} is a constant expression is checked exactly once, at compile time, and
then at runtime, the implementation may assume that every time control flow
passes through the definition of \tcode{r}, if the initialization \emph{were}
evaluated, it would bind \tcode{r} to the same object as if \tcode{o} were
nominated, and if this assumption does not hold, the behavior is undefined.

The advantages of this approach are:
\begin{itemize}
\item The programmer never needs to restructure their code in order to convince
the compiler not to reject it for having a potentially dangling
\tcode{constexpr} reference.
\item The feature as a whole might end up being easier to explain to a learner.
In fact, this approach may be the one that most closely approximates the
one-sentence title of this paper. The restrictions imposed by approaches 3a and
3b would not need to be taught since they would not be present at all, and UB
would not result as long as the programmer writes ``reasonable'' code.
\item This approach is the most consistent with the general philosophy that
\tcode{constexpr} has been moving toward, i.e., that you only get an ill-formed
program when undefined behavior is \emph{actually reached} during constant
evaluation.
\end{itemize}
The main disadvantages are:
\begin{itemize}
\item The specification and implementation difficulties are unknown.
\item In current C++, programmers do not need to worry about the lifetime of an
object that is bound to a \tcode{constexpr} reference. Thus, an lvalue-to-rvalue
conversion on the name of the reference will always have well-defined behavior,
just like an lvalue-to-rvalue conversion on the name of a \tcode{constexpr}
object. Under this approach, programmers who have never needed to worry about
the lifetime of referents of \tcode{constexpr} references may write code that
has undefined behavior because they are used to not worrying about the lifetime.
\end{itemize}
The approach described above is suboption 3c. Suboption 3d is very similar, but
also allows the compiler to reject \tcode{f2} as a matter of quality of
implementation (QoI). To be more precise, the implementation can reject the
program if it determines that there is any combination of function argument
values and global state that would result in a \tcode{constexpr} reference to
one of that function's automatic variables escaping from the function, but if
the implementation doesn't know whether or not this is possible, then the
program merely has UB at runtime if such a dangling reference is accessed (just
like if the reference were not \tcode{constexpr}). In jargon, we say that such
a function is ``unspeciformed'' (i.e., it's unspecified whether the
implementation treats it as if it were ill formed). The main advantages of
suboption 3d are:
\begin{itemize}
\item As long as the programmer can prove that no possible values of function
arguments can result in a \tcode{constexpr} reference to a local variable
escaping from the function, they won't need to worry about the compiler
rejecting the code.
\item Some unsafe constructs may be diagnosed, which makes this approach safer
than suboption 3c.
\end{itemize}
The main disadvantages are:
\begin{itemize}
\item As with suboption 3c, the specification and implementation difficulties of
requiring a single abstract machine initialization of a \tcode{constexpr}
reference with automatic storage duration are unknown.
\item Unspeciformed behavior is novel. CWG has not yet had the opportunity to
weigh in on any potential specification or implementation issues that would
arise as a result of unspeciformed behavior being in the standard.
\item This approach has false negatives; in cases where the programmer has
written defective code because they assumed that \tcode{constexpr} references
never dangle, the compiler will not always be able to detect this defective
assumption.
\item It might be that the program logic prevents the function from ever being
called with arguments that result in escape, but this cannot be determined by
the compiler, because those function argument values could e.g. come from
calling a function whose definition is in another translation unit. Thus, in a
sense, this approach has false positives: the compiler would be permitted to
diagnose the function definition because it has determined that there is a set
of argument values that results in escape, even though it doesn't know whether
any set of inputs to the program could result in those argument values.
\item Unspeciformed behavior has a teachability burden because the same code may
be accepted by some implementations and rejected by others; learners will wonder
what the ``right'' answer is.
\end{itemize}

\subsubsection{Conclusion}
We consider suboption 3a to be the most preferable because it is easy to
specify, we believe there are no implementation difficulties, and it has not
been shown that any other suboption is more teachable, with the possible
exception of 3c, which suffers from unknown specification and implementation
difficulty. To be more precise, suboption 3a is the simplest approache that
achieves the design goals of symbolic addressing (and thus, \tcode{constexpr}
structured bindings with automatic storage duration).

\section{Wording for Option 3a (symbolic addressing with the same-function restriction)}

\rSec1[basic.def.odr]{One-definition rule}%

\ednote{Modify p5 as follows:}

\label{term.odr.use}%
A variable is named by an expression
if the expression is an \grammarterm{id-expression} that denotes it.
A variable \tcode{x} that is named by a
potentially-evaluated expression \changed{$E$}{$N$}\added{ that appears at a point $P$}
is \defnx{odr-used}{odr-use} by \changed{$E$}{$N$} unless
\added{$N$ is an
element of the set of potential results of an expression $E$, where
}
\begin{removedblock}
\begin{itemize}
    \item
    \tcode{x} is a reference that is
    usable in constant expressions \iref{expr.const}, or
    \item
    \tcode{x} is a variable of non-reference type that is
    usable in constant expressions and has no mutable subobjects, and
    $E$ is an element of the set of potential results of an expression
    of non-volatile-qualified non-class type
    to which the lvalue-to-rvalue conversion \iref{conv.lval} is applied, or
    \item
    \tcode{x} is a variable of non-reference type, and
    $E$ is an element of the set of potential results
    of a discarded-value expression \iref{expr.context}
    to which the lvalue-to-rvalue conversion is not applied.
\end{itemize}
\end{removedblock}

\draftnote{Ideally, \tcode{x} should be allowed to have mutable subobjects as
long as we don't touch the mutable parts of \tcode{x}. This would probably only
require slightly more complex wording, but isn't in scope for this paper because
the status quo also doesn't allow it.}

\begin{addedblock}
\begin{itemize}
\item
\tcode{x} is a reference that is usable in constant expressions at $P$,
\item
\tcode{x} is a non-volatile object that is usable in constant expressions at $P$and has no mutable subobjects, and $E$ is a class member access expression ([expr.ref]) naming a non-static data member of reference type,
\item
\tcode{x} is a non-volatile object that is usable in constant expressions at $P$and has no mutable subobjects, and the lvalue-to-rvalue
conversion (\iref{conv.lval}) is applied to $E$, or
\item
$E$ is a discarded-value expression (\iref{expr.context}) to which the lvalue-to-rvalue
conversion is not applied.
\end{itemize}
\begin{example}
\begin{codeblock}
int f(int);
int g(int&);
struct A {
    int x;
};
struct B {
    int& r;
};
int h(bool cond) {
    constexpr A a = {1};
    constexpr volatile A& r = a;  // odr-uses \tcode{a}
    int _ = f(cond ? a.x : r.x);  // does not odr-use \tcode{a} or \tcode{r}
    int x, y;
    constexpr B b1 = {x}, b2 = {y};  // odr-uses \tcode{x} and \tcode{y}
    int _ = g(cond ? b1.r : b2.r);  // does not odr-use \tcode{b1} or \tcode{b2}
    int _ = (cond ? x : y), 0;  // does not odr-use \tcode{x} or \tcode{y}
    return [] {
        return b1.r;  // error: \tcode{b1} is odr-used here because the object
                      // referred to by \tcode{b1.r} is not constexpr-referenceable
                      // here
    };
}
\end{codeblock}
\end{example}
\end{addedblock}

\rSec3[basic.start.static]{Static initialization}

\draftnote{The constant initialization of a variable implicitly
includes the constant initialization of any temporary objects whose lifetimes
are extended to that of the variable. All references to constant initialization
from elsewhere in the standard currently refer only to variables with constant
initialization. Removing the words "or temporary object" from this paragraph
simplifies the wording elsewhere by avoiding the need to define when an object
(as opposed to variable) is constant-initialized.}

\ednote{Modify p2 as follows:}

\pnum
\indextext{initialization!constant}%
\defnx{Constant initialization}{constant initialization} is performed
if a variable \removed{or temporary object} with static or thread storage duration
is constant-initialized\iref{expr.const}.
\indextext{initialization!zero-initialization}%
If constant initialization is not performed, a variable with static
storage duration\iref{basic.stc.static} or thread storage
duration\iref{basic.stc.thread} is zero-initialized\iref{dcl.init}.
Together, zero-initialization and constant initialization are called
\defnadj{static}{initialization};
all other initialization is \defnadj{dynamic}{initialization}.
All static initialization strongly happens before\iref{intro.races}
any dynamic initialization.
\begin{note}
    The dynamic initialization of non-block variables is described
    in \iref{basic.start.dynamic}; that of static block variables is described
    in \iref{stmt.dcl}.
\end{note}

\rSec1[expr.const]{Constant expressions}
\indextext{expression!constant}

\draftnote{\paper{P0784R7} abolished the previous restriction that constexpr
constructors of non-literal class types may not be invoked during constant
evaluation. The current wording of [expr.const]/2 still contains a special
exception that allows a variable to be considered constant-initialized even
though the initialization would invoke such a constructor; that wording is
unnecessary since \paper{P0784R7} was accepted.}

\draftnote{A structured binding is a named lvalue, but is not a
reference in the non-tuple-like cases; therefore, the current rules regarding references that are not
usable in constant expressions ([expr.const]/8) do not always apply to structured
bindings. The intent of the below wording is that structured bindings should be
subject to the same restrictions during constant evaluation that would apply if
they were references.
}

% https://eel.is/c++draft/dcl.struct.bind#4

\draftnote{The definition of "constexpr-referenceable" below is
written under the assumption that temporary objects are considered to have the
storage duration described in \paper{CWG1634}, namely, that a temporary object whose
lifetime is extended inherits the storage duration of the reference that is
bound to it, and any other temproary object has a distinct storage duration.
}

\pnum
Certain contexts require expressions that satisfy additional
requirements as detailed in this subclause; other contexts have different
semantics depending on whether or not an expression satisfies these requirements.
Expressions that satisfy these requirements,
assuming that copy elision\iref{class.copy.elision} is not performed,
are called
\indexdefn{expression!constant}%
\defnx{constant expressions}{constant expression}.
\begin{note}
    Constant expressions can be evaluated
    during translation.
\end{note}

\begin{bnf}
    \nontermdef{constant-expression}\br
    conditional-expression
\end{bnf}

\ednote{Insert a paragraph after p1:}

\begin{addedblock}
The \defn{constituent values} of an object $o$ are the value of $o$ if it has scalar type and the values of
any of $o$’s subobjects of scalar type, other than inactive union members and subobjects
thereof. The \defn{constituent references} of an object $o$ are the non-static data members of
reference type of $o$ and of any of $o$’s subobjects that are neither inactive union members nor
subobjects thereof.
\end{addedblock}

\ednote{Insert a paragraph after p1:}

\begin{addedblock}
The constituent values and constituent references of a variable \tcode{x} are defined as follows:
\begin{itemize}
\item If \tcode{x} declares an object, the constituent values and references of that object are
constituent values and references of \tcode{x}.
\item If \tcode{x} declares a reference, that reference is a constituent reference of \tcode{x}.
\end{itemize}
For any constituent reference \tcode{r} of a variable \tcode{x}, if \tcode{r} is bound to a temporary object or
subobject thereof whose lifetime is extended to that of \tcode{r}, the constituent values and
references of that temporary object are also constituent values and references of \tcode{x}. This rule
applies recursively.
\end{addedblock}

\ednote{Insert a paragraph after p1:}

\begin{addedblock}
An object $o$ is \defn{constexpr-referenceable} from a point $P$ if
\begin{itemize}
\item $o$ has static storage duration, or
\item $o$ has automatic storage duration, and, letting \tcode{v} denote the variable corresponding to
$o$’s complete object, or the variable to whose lifetime that of $o$ is extended, \tcode{v}’s smallest
enclosing non-block scope and $P$’s smallest enclosing non-block scope are the same function parameter scope.
\end{itemize}
\begin{example}
\begin{codeblock}
struct A {
    int m;
    const int& r;
};
void f() {
    static int sx;
    thread_local int tx;
    int ax;
    A aa = {1, 2};
    static A sa = {3, 4};
    // The objects \tcode{sx}, \tcode{ax}, and \tcode{aa.m}, \tcode{sa.m}, and
    // the temporaries to which \tcode{aa.r} and \tcode{sa.r} are bound, are
    // constexpr-referenceable.
    auto lambda = [] {
        int ay;
        // The objects \tcode{sx}, \tcode{sa.m}, and \tcode{ay}, and the
        // temporary to which \tcode{sa.r} is bound, are constexpr-referenceable.
    };
}
\end{codeblock}
\end{example}
\end{addedblock}

\ednote{Insert a paragraph after p1:}

\begin{addedblock}
An object or reference \tcode{x} is \defn{constexpr-representable} at a point $P$ if, for each constituent value
of \tcode{x} that points to or past an object $o$, and for each constituent reference of \tcode{x} that refers to an
object $o$, $o$ is constexpr-referenceable from $P$.
\end{addedblock}

\ednote{Modify p2 as follows:}

\pnum
A variable \removed{or temporary object} \changed{\tcode{o}}\tcode{v}} is \defn{constant-initialized} if
\begin{itemize}
\item
either it has an initializer or
its default-initialization results in some initialization being performed, \removed{and}
\item
the full-expression of its initialization is a constant expression
when interpreted as a \grammarterm{constant-expression}\removed{,
except that if \tcode{o} is an object,
that full-expression
may also invoke constexpr constructors
for \tcode{o} and its subobjects
even if those objects are of non-literal class types.}
\begin{note}
    \removed{Such a class can have a non-trivial destructor.}
    Within this evaluation
    \tcode{std::is_constant_evaluated()} \iref{meta.const.eval}
    returns \keyword{true}.
\end{note}\added{, and}
\begin{addedblock}
\item immediately after the initializing declaration of \tcode{v}, the object or reference declared by \tcode{v} is constexpr-representable, and if \tcode{v} has static or thread storage duration, the object or reference declared by \tcode{v} is constexpr-representable at a point that follows the initializing declaration of \tcode{v} and whose immediate scope is a namespace scope.
\end{addedblock}
\end{itemize}
\begin{addedblock}
\begin{example}
\begin{codeblock}
void f() {
    int ax;                      // \tcode{ax} is not constant-initialized
    thread_local int tx;         // \tcode{tx} is constant-initialized
    static int sx;               // \tcode{sx} is constant-initialized
    static int& rss = sx;        // \tcode{rss} is constant-initialized
    static int& rst = tx;        // \tcode{rst} is not constant-initialized
    static int& rsa = ax;        // \tcode{rsa} is not constant-initialized
    thread_local int& rts = sx;  // \tcode{rts} is constant-initialized
    thread_local int& rtt = tx;  // \tcode{rtt} is not constant-initialized
    thread_local int& rta = ax;  // \tcode{rta} is not constant-initialized
    int& ras = sx;               // \tcode{ras} is constant-initialized
    int& rat = tx;               // \tcode{rat} is not constant-initialized
    int& raa = ax;               // \tcode{raa} is constant-initialized
}
\end{codeblock}
\end{example}
\end{addedblock}

\pnum
A variable is \defn{potentially-constant} if
it is constexpr or
it has reference or non-volatile const-qualified integral or enumeration type.

\ednote{Modify p4 as follows:}

\pnum
A constant-initialized potentially-constant variable $V$ is
\defn{usable in constant expressions} at a point $P$ if
$V$'s initializing declaration $D$ is reachable from $P$ and
\begin{itemize}
    \item $V$ is constexpr,
    \item $V$ is not initialized to a TU-local value, or
    \item $P$ is in the same translation unit as $D$.
\end{itemize}
An object or reference \added{$x$} is \defn{usable in constant expressions} \added{at point $P$} if it is
\begin{itemize}
    \begin{removedblock}
    \item a variable that is usable in constant expressions, or
    \end{removedblock}
    \item a template parameter object \iref{temp.param}\added{ or subobject thereof}, or
    \item a string literal object \iref{lex.string}\added{ or subobject thereof}, or
    \begin{removedblock}
    \item a temporary object of non-volatile const-qualified literal type
    whose lifetime is extended \iref{class.temporary}
    to that of a variable that is usable in constant expressions, or
    \item a non-mutable subobject or reference member of any of the above.
    \end{removedblock}
    \begin{addedblock}
    \item one of the following:
    \begin{itemize}
        \item a variable that is usable in constant expressions at $P$,
        \item a temporary object of non-volatile const-qualified literal type
        whose lifetime is extended (\iref{class.temporary})
        to that of a variable that is usable in constant expressions at $P$,
        \item a non-mutable subobject of any of the above, or
        \item a reference member of any of the above
    \end{itemize}
    that is constexpr-representable at $P$.
    \end{addedblock}
\end{itemize}
\begin{addedblock}
\begin{example}
\begin{codeblock}
struct A {
    int* const & r;
};
void f(int x) {
    constexpr A a = {&x};
    static_assert(a.r == &x);  // OK
    [&] {
        static_assert(a.r != nullptr);  // error: \tcode{a.r} is not usable in
                                        // constant expressions at this point
    }();
}
\end{codeblock}
\end{example}
\end{addedblock}

\ednote{Add after p8:}

\begin{addedblock}
For the purposes of determining whether an expression is a core constant expression, the
evaluation of an \grammarterm{id-expression} that names a structured binding \tcode{v} (\iref{dcl.struct.bind}) has the
following semantics:
\begin{itemize}
\item If \tcode{v} is an lvalue referring to the object bound to an invented reference \tcode{r}, the
behavior is as if \tcode{r} were nominated.
\item Otherwise, if \tcode{v} names an array member or class member, the behavior is that of
evaluating \tcode{$e$[$i$]} or \tcode{$e$.$m$}, respectively, where \tcode{$e$} is the name of the variable initialized
from the initializer of the structured binding declaration, and \tcode{$i$} is the index of the
element referred to or \tcode{$m$} is the name of the member referred to by \tcode{v}, respectively.
\end{itemize}
\begin{example}
\begin{codeblock}
#include <tuple>
struct A {
    virtual int g() const;
};
void f(std::tuple<A&> a) {
    auto [r] = a;
    static_assert(r.g() >= 0);  // error: dynamic type is constexpr-unknown
}
\end{codeblock}
\end{example}
\end{addedblock}

\ednote{Modify p13 as follows:}

\pnum
A \defnadj{constant}{expression} is either
a glvalue core constant expression that refers to
\changed{an entity that is a permitted result of a constant expression (as defined below)}{an object or a non-immediate function}, or
a prvalue core constant expression whose value
satisfies the following constraints:
\begin{itemize}
    \begin{removedblock}
    \item
    if the value is an object of class type,
    each non-static data member of reference type refers to
    an entity that is a permitted result of a constant expression,

    \item
    if the value is an object of scalar type,
    it does not have an indeterminate value\iref{basic.indet},

    \item
    if the value is of pointer type, it contains
    the address of an object with static storage duration,
    the address past the end of such an object\iref{expr.add},
    the address of a non-immediate function,
    or a null pointer value,

    \item
    if the value is of pointer-to-member-function type,
    it does not designate an immediate function, and

    \item
    if the value is an object of class or array type,
    each subobject satisfies these constraints for the value.
    \end{removedblock}
    \begin{addedblock}
    \item each constituent reference refers to an object or a non-immediate function,
    \item no constituent value of scalar type is an indeterminate value ([basic.indet]),
    \item no constituent value of pointer type is a pointer to an immediate function or an invalid pointer value (\iref{basic.compound}), and
    \item no constituent value of pointer-to-member type designates an immediate function.
    \end{addedblock}
\end{itemize}
\begin{removedblock}
An entity is a
\defnx{permitted result of a constant expression}{constant expression!permitted result of}
if it is an
object with static storage duration that either is not a temporary object or is
a temporary object whose value satisfies the above constraints, or if
it is a non-immediate function.
\end{removedblock}
\begin{note}
    A glvalue core constant expression
    that either refers to or points to an unspecified object
    is not a constant expression.
\end{note}


\rSec1[dcl.dcl]{Declarations}%
\rSec2[dcl.pre]{Preamble}

\ednote{Change p6 as follows:}

\pnum
A \grammarterm{simple-declaration} with an \grammarterm{identifier-list} is called
a \defn{structured binding declaration} \iref{dcl.struct.bind}.
Each \grammarterm{decl-specifier} in the \grammarterm{decl-specifier-seq}
shall be
\added{\tcode{constexpr}, \tcode{constinit},}
\tcode{static},
\tcode{thread_local},
\tcode{auto} \iref{dcl.spec.auto}, or
a \grammarterm{cv-qualifier}.
\begin{example}
    \begin{codeblock}
        template<class T> concept C = true;
        C auto [x, y] = std::pair{1, 2};    // error: constrained \grammarterm{placeholder-type-specifier}
        // not permitted for structured bindings
    \end{codeblock}
\end{example}

\rSec1[dcl.struct.bind]{Structured binding declarations}%
%\indextext{structured binding declaration}%
%\indextext{declaration!structured binding|see{structured binding declaration}}%

\ednote{Change p1 as follows:}

\pnum
A structured binding declaration introduces the \grammarterm{identifier}{s}
$\tcode{v}_0$, $\tcode{v}_1$, $\tcode{v}_2, \dotsc$
of the
\grammarterm{identifier-list} as names
of \defn{structured binding}{s}.
Let \cv{} denote the \grammarterm{cv-qualifier}{s} in
the \grammarterm{decl-specifier-seq} and
\placeholder{S} consist of \changed{the \grammarterm{storage-class-specifier}{s} of
the \grammarterm{decl-specifier-seq} (if any)}
{each \grammarterm{decl-specifier} of the \grammarterm{decl-specifier-seq} that
is \tcode{constexpr}, \tcode{constinit}, or a
\grammarterm{storage-class-specifier}}.
A \cv{} that includes \tcode{volatile} is deprecated;
see [depr.volatile.type].
First, a variable with a unique name \exposid{e} is introduced. If the
\grammarterm{assignment-expression} in the \grammarterm{initializer}
has array type \cvqual{cv1} \tcode{A} and no \grammarterm{ref-qualifier} is present,
\exposid{e} is defined by
\begin{ncbnf}
    \opt{attribute-specifier-seq} \placeholder{S} \cv{} \terminal{A} \exposid{e} \terminal{;}
\end{ncbnf}
and each element is copy-initialized or direct-initialized
from the corresponding element of the \grammarterm{assignment-expression} as specified
by the form of the \grammarterm{initializer}.
Otherwise, \exposid{e}
is defined as-if by
\begin{ncbnf}
    \opt{attribute-specifier-seq} decl-specifier-seq \opt{ref-qualifier} \exposid{e} initializer \terminal{;}
\end{ncbnf}
where
the declaration is never interpreted as a function declaration and
the parts of the declaration other than the \grammarterm{declarator-id} are taken
from the corresponding structured binding declaration.
The type of the \grammarterm{id-expression}
\exposid{e} is called \tcode{E}.
\begin{note}
    \tcode{E} is never a reference type \iref{expr.prop}.
\end{note}

\pnum
If the \grammarterm{initializer} refers to
one of the names introduced by the structured binding declaration,
the program is ill-formed.

\pnum
If \tcode{E} is an array type with element type \tcode{T}, the number
of elements in the \grammarterm{identifier-list} shall be equal to the
number of elements of \tcode{E}. Each $\tcode{v}_i$ is the name of an
lvalue that refers to the element $i$ of the array and whose type
is \tcode{T}; the referenced type is \tcode{T}.
\begin{note}
    The top-level cv-qualifiers of \tcode{T} are \cv.
\end{note}
\begin{example}
    \begin{codeblock}
        auto f() -> int(&)[2];
        auto [ x, y ] = f();            // \tcode{x} and \tcode{y} refer to elements in a copy of the array return value
        auto& [ xr, yr ] = f();         // \tcode{xr} and \tcode{yr} refer to elements in the array referred to by \tcode{f}'s return value
    \end{codeblock}
\end{example}


\rSec2[dcl.constexpr]{The \keyword{constexpr} and \keyword{consteval} specifiers}%
\indextext{specifier!\idxcode{constexpr}}
\indextext{specifier!\idxcode{consteval}}

\ednote{Change p1 as follows:}

\pnum
The \keyword{constexpr} specifier shall be applied only to
the definition of a variable or variable template\added{, a structured binding declaration, } or
the declaration of a function or function template.
The \keyword{consteval} specifier shall be applied only to
the declaration of a function or function template.
A function or static data member
declared with the \keyword{constexpr} or \keyword{consteval} specifier
is implicitly an inline function or variable \iref{dcl.inline}.
If any declaration of a function or function template has
a \keyword{constexpr} or \keyword{consteval} specifier,
then all its declarations shall contain the same specifier.

[...]

\ednote{Modify p6 as follows:}

\pnum
A \keyword{constexpr} specifier used in an object declaration
declares the object as const.
Such an object
shall have literal type and
shall be initialized.
In any \keyword{constexpr} variable declaration,
the full-expression of the initialization
shall be a constant expression \iref{expr.const}.
A \keyword{constexpr} variable that is an object,
as well as any temporary to which a \keyword{constexpr} reference is bound,
shall have constant destruction.
\added{
Let $x$ denote the object or reference declared by the declaration of a
\tcode{constexpr} variable \tcode{v}.
Immediately after the initializing declaration of \tcode{v},
$x$ shall be constexpr-representable.
If $x$ has static or thread storage duration, $x$ shall be
constexpr-representable at a point following the initializing declaration of
\tcode{v} whose immediate scope is a namespace scope.
}

\begin{example}
\begin{codeblock}
struct pixel {
    int x, y;
};
constexpr pixel ur = { 1294, 1024 };    // OK
constexpr pixel origin;                 // error: initializer missing
\end{codeblock}
\begin{addedblock}
\begin{codeblock}
namespace N {
void f() {
    int x;
    constexpr int& ar = x;         // OK
    static constexpr int& sr = x;  // error: \tcode{x} is not
                                   // constexpr-representable at the point
                                   // indicated below
}
// immediate scope here is that of \tcode{N}
}
\end{codeblock}
\end{addedblock}
\end{example}

\rSec2[dcl.constexpr]{The \keyword{constinit} specifier}

\emph{Drafting note:} Unlike in [dcl.constexpr], we don't need an explicit rule
about the object or reference being constexpr-representable in this section,
because the restriction added to [expr.const]/2 will cause the variable to have
dynamic initialization if the object or reference is not
constexpr-representable.

\ednote{Modify p1 as follows:}

The \keyword{constinit} specifier shall be applied only to a declaration of a
variable with static or thread storage duration \added{or to a structured
binding declaration (\iref{dcl.struct.bind})}. If the specifier is applied to
any declaration of a variable, it shall be applied to the initializing
declaration. No diagnostic is required if no \keyword{constinit} declaration is
reachable at the point of the initializing declaration.

\rSec2[temp.arg.nontype]{Template non-type arguments}

\ednote{Modify [temp.arg.nontype]/p1 as follows:}

\added{%
A \grammarterm{template-argument} for a non-type  \grammarterm{template-parameter} with declared type \tcode{T} shall be such
that the invented declaration%
}
\begin{addedblock}
\begin{codeblock}
    T x = @\grammarterm{template-argument}@ ;
\end{codeblock}
\end{addedblock}
\added{%
satisfies the semantic constraints for the definition of a \tcode{constexpr} variable with static
storage duration [dcl.constexpr].
}%
If \removed{the type} \tcode{T} \removed{of a \grammarterm{template-parameter} \iref{temp.param}}
contains a placeholder type (\iref{dcl.spec.auto})
or a placeholder for a deduced class type (\iref{dcl.type.class.deduct}),
the type of the parameter is \changed{the type deduced for the variable \tcode{x} in the invented declaration}{deduced from the above declaration.}
\begin{removedblock}
\begin{codeblock}
    T x = @\grammarterm{template-argument}@ ;
\end{codeblock}
\end{removedblock}
\changed{If a deduced parameter type}{If the parameter type thus deduced} is not permitted
for a \grammarterm{template-parameter} declaration \iref{temp.param},
the program is ill-formed.

\ednote{Add after [temp.arg.nontype]/p6:}

\begin{addedblock}
\begin{example}
\begin{codeblock}
template <int& r> class A;
extern int x;
A<x> a;  // OK
void f(int p) {
    constexpr int& r = p;  // OK
    A<r> a;                // error: a \tcode{static constexpr int\&} variable
                           // cannot be initialized to refer to \tcode{p} here
}
\end{codeblock}
\end{example}
\end{addedblock}

\section{Feature test macros}

\ednote{ In \tcode{[tab:cpp.predefined.ft]}, bump \tcode{__cpp_structured_bindings} to the date of adoption}.


\section{Acknowledgments}

We would like to thank Bloomberg for sponsoring this work.
Thanks to Nina Dinka Ranns, Pablo Halpern, and Joshua Berne for their feedback.

Thanks to Richard Smith for the original discussion of possible solutions on the Core reflector and for feedback on the wording.

Thanks to Nicolas Lesser for the original work on \paper{P1481R0}.

Thanks to Daisy Hollman for input on the issue of lambdas that reference
variables from their enclosing functions.

\section{References}

\renewcommand{\section}[2]{}%
\bibliographystyle{plain}
\bibliography{wg21, extra}

\begin{thebibliography}{9}

\bibitem[N4885]{N4885}
Thomas Köppe
\emph{Working Draft, Standard for Programming Language C++}\newline
\url{https://wg21.link/N4885}


\end{thebibliography}

\end{document}

%%%% example foo6 on godbolt is usable as a pr-value in a constant expression
% [expr.const] 5.9
% special case structure bindings to say they are evaluated in a constant expression context.
% Modify https://eel.is/c++draft/dcl.pre#6
